---
title: "TopicAnalysis_pipeline"
author: "Maria D Hernandez Limon - MDHL"
date: "2026-04-27"
output:
  pdf_document: default
  html_document: default
---

#data 
FL <- Free Living prokaryotes 0.2-1.6um
--SSF is the chloroplast from FL annoated with PRv5
PA <- Particle Associated prokaryotes >1.6um
--LSF is the chlororplast from PA annoated with PRv5

#Background
Text:
https://academic.oup.com/biostatistics/article/20/4/599/5032578
https://www.tidytextmining.com/topicmodeling.html
Examples:
https://github.com/lasy/alto/blob/master/vignettes/microbiome-demo.Rmd
https://lasy.github.io/alto/articles/lda.html

really helpful descriptions:
https://www.analyticsvidhya.com/blog/2021/06/part-2-topic-modeling-and-latent-dirichlet-allocation-lda-using-gensim-and-sklearn/

#1. Load libraries and set-up
```{r}
#install.packages("devtools")
#devtools::install_github("lasy/alto")

#install.packages("pak")
#pak::pkg_install("lasy/alto")
  
library(alto) #breaks for me if I try to use anr R greater rgab 4.4.1
library(topicmodels)
library(betareg)

library(dplyr)
library(purrr)
library(magrittr)
library(ggplot2)
library(tibble)
library(stringr)
library(tidyr)
library(reshape)
library(Hmisc)
library(cluster)
library(phyloseq)
library(corrplot)
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(car)
library(flextable)
library(cowplot)
library(ggpubr)
library(RColorBrewer) 
library(grid) 

library(ggrepel)
library(ggforce)
```


##colors
```{r}
phyla_colors <- c(
  'Pseudomonadota'='#bae342',        # was Proteobacteria
  'Cyanobacteriota'='#1c5b5a',       # was Cyanobacteria
  'Actinomycetota'='#860967',        # was Actinobacteriota
  'Bacillota'='blue',                # was Firmicutes
  'Chloroflexota'='#888bce',         # was Chloroflexi
  'Thermodesulfobacteriota'='salmon',# was Desulfobacterota
  'Thermoproteota'='orange',         # was Crenarchaeota
  'Verrucomicrobiota'='#dc5dd8',
  'Bacteroidota'='#f92e5d',
  'Planctomycetota'='#c0583d',
  'Acidobacteriota'='#514e72',
  'Nitrospirota'='#6edb9f',
  'Bdellovibrionota'='yellow',
  'Armatimonadota'='pink',
  'Myxococcota'='lightblue',
  'unclass_Bacteria'='maroon',
  'unclass_Archaea'='orange4',
  'Dependentiae'="red4", 
  'Chlamydiota'="skyblue", 
  'Gemmatimonadota'="green",
  'Other'='gray'
)

class_colors <- c(
  "Bacillariophyceae" = "#45AA45",   "Cryptophyceae" = "#FF800D",   "Prymnesiophyceae" = "#9467bd",
  "Chlorophyceae" = "green",      "Eustigmatophyceae" = "#FFD0C9",   "Chrysophyceae" = "cyan",
  "Trebouxiophyceae" = "#FEE11C",   "Diatomeae_X" = "#FF0073",     "Pavlovophyceae" = "navy",
  "Dictyochophyceae" = "#90164F",   "Zygnemophyceae" = "#F6A5FD",  "Mamiellophyceae" = "red",
  "Chlorodendrophyceae" = "olivedrab", "Euglenida" = "#9B5D00",      "Klebsormidiophyceae" = "#0D75F9",
  "Xanthophyceae" = "#6C16A0",      "Pelagophyceae" = "#C2E1F1",   "Bangiophyceae" = "#BCEF00",
  "Mediophyceae" = "#DEE99F",       "unclass_plas" = "black",      "Other" = "gray"
)

sf_colors <- c(
  "FL_Prok" = "#1f78b4",   # blue
  "PA_Prok" = "#b15928",   # brown
  "SSF_Chloro" = "#33a02c",  # light green
  "LSF_Chloro" = "#00441b"   # dark green
)

sf_colors2 <- c(
  "FL" = "#1f78b4",   # blue
  "PA" = "#b15928",   # brown
  "SSF" = "#33a02c",  # light green
  "LSF" = "#00441b",   # dark green
  "Env" ="red"
)

year_colors=c('2012'='red','2013'='#A6CEE3','2014'='#B2DF8A','2015'="brown",
                                 '2016'="violet",'2017'="purple",'2018'="blue",'2019'="black", "2022"="salmon")

lake_shapes=c("SU"=21,"MI"=22, "HU"=23,"ER"=24,"ON"=25)

year_shapes=c('2012'=0,'2013'=1,'2014'=2,'2015'=3,'2016'=4,'2017'=5,'2018'=6,'2019'=7, "2022"=21)

depth_colors=c( "Top"="#90CAF9","Mid"="#1E88E5","Bot"="#0D47A1",
                "EPI"="#F1B02FFF","DCL"="#59A55DFF","HYP"="#5E3194FF","TRM"="gray")

depth_colors_w=c('Iso_Top'="#4CBED9FF","Iso_Mid"="#4CBED9FF","Iso_Bot"="#4CBED9FF",
                 "InvS_EPI"="#C34582FF","InvS_TRM"="gray","InvS_HYP"="#9870B9FF",
                "Strat_EPI"="#F1B02FFF","Iso_EPI"="red","Strat_TRM"="gray",
                "Strat_DCL"="#59A55DFF","Strat_HYP"="#5E3194FF","Spring"="#7DA7EAFF")

iso_colors=c("Top"="#90CAF9","Mid"="#1E88E5","Bot"="#0D47A1")
invs_colors=c("EPI"="#C34582FF","TRM"="gray","HYP"="#9870B9FF")
norm_colors=c("EPI"="#F1B02FFF","DCL"="#4DAF4A","TRM"="gray","HYP"="#5E3194FF")


depth2_colors=c('Iso_Top'="#90CAF9","Iso_Mid"="#1E88E5","Iso_Bot"="#0D47A1",
                "Strat_EPI"="#F1B02FFF","Strat_DCL"="#59A55DFF","Strat_HYP"="#5E3194FF",
                "Strat_TRM"="gray","Iso_EPI"="red",
                "InvS_EPI"="#C34582FF","InvS_TRM"="gray","InvS_HYP"="#9870B9FF")

strat_colors=c("Iso"="#ffd700","InvS"="#80B1D3","Strat"="#FB8072","Norm"="#FB8072")

lake_colors=c("SU"="#92809E","MI"="#8cabea", "HU"="#64baaa","ER"="#FFC86A","ON"="#ee7061")

season_colors=c("Summer"="#F69541","Spring"="#7DA7EAFF","Winter"="#1D4497FF")

```

##data dirs
```{r setup, include=TRUE}
knitr::opts_knit$set("/Volumes/mlcoleman/project/greatlakes/combined_1618s_mdh/Analysis_with_both/LDA/to_share/")
```

```{r}
FL_data<-readRDS("/Volumes/mlcoleman/project/greatlakes/combined_1618s_mdh/Manuscripts/PROK/test/data/ps1_FL.rds")
PA_data<-readRDS("/Volumes/mlcoleman/project/greatlakes/combined_1618s_mdh/Manuscripts/PROK/test/data/ps1.rds")
SSF_data<-readRDS("/Volumes/mlcoleman/project/greatlakes/combined_1618s_mdh/PC_FL/EUK_analysis/R_code/output/ps1_SSF.rds")
LSF_data<-readRDS("/Volumes/mlcoleman/project/greatlakes/combined_1618s_mdh/MPBio_PA/EUK_analysis/R_code/output/ps1_LSF.rds")
```

#____________
In this section I will introduce the functions that get used again for the different biological blocks

```{r}
#colors
library(RColorBrewer)

# Get the 10-color Paired palette
paired_colors <- brewer.pal(10, "Paired")

# Create FL colors
FL_colors <- paired_colors
names(FL_colors) <- paste0("FL_", seq_along(FL_colors))

# Create PA colors (reusing the same palette)
PA_colors <- paired_colors
names(PA_colors) <- paste0("PA_", seq_along(PA_colors))

# Combine into one named vector
prok_scs <- c(FL_colors, PA_colors)

prok_scs

```

#**FL**
#2. Prep data 
```{r}
process_phyloseq <- function(ps) {
  library(phyloseq)
  
  # Extract the three tables
  if (taxa_are_rows(ps)) {
    asv <- as.data.frame(t(otu_table(ps)))
  } else {
    asv <- as.data.frame(otu_table(ps))
  }
  
  tax  <- as.data.frame(tax_table(ps))
  meta <- as.data.frame(sample_data(ps))
  
  # Hand off to process_tables so we can clean if needed 
  return(process_tables(asv, tax, meta))
}

process_tables <- function(asv, tax, meta) {
  library(dplyr)
  
  tax$ASV <- rownames(tax)
  
  #we need asvs to be unique so we add a numer
  tax_dict <- tax %>%
    mutate(
      ASV_name = paste0(
        ifelse(is.na(Phylum), "[unknown Phylum]", Phylum),
        " (",
        ifelse(is.na(Order), "?", Order),
        ") ",
        row_number()
      )
    ) %>%
    ungroup()
  
  tax_dict$ASV <- factor(tax_dict$ASV, levels = colnames(asv))
  tax_dict <- tax_dict[order(tax_dict$ASV), ]
  
  colnames(asv) <- tax_dict$ASV_name
  asv <- as.data.frame(asv) %>% mutate(across(everything(), as.integer))
  
  return(list(
    counts   = as.data.frame(asv),
    tax_dict = tax_dict,
    meta     = meta
  ))
}
```

```{r}
#this inlcudes counts, tax_dict, meta
FL_results<-process_phyloseq(FL_data)

#save the counts tables you will need for perplexity
write.csv(FL_results$counts, "output/tables/runFL/FL_counts.csv")

#free up some memoery we don't need the FL_data anymore and we can call it in if we do 
rm(FL_data)
gc()  

```

#3. Run LDA
Our first step is to run the lda models for varying number of topics, i.e. from 1 to 20 topics.
For this, we use the `run_lda_models` function. This can take a while on a personal computer but we will save the runs so future runs won't take long.

### run LDA models 
```{r run-lda-models}
lda_varying_params_lists <-  list()
for (k in 1:20) {lda_varying_params_lists[[paste0("k",k)]] <- list(k = k)}

lda_models  <- 
  run_lda_models(
    data = FL_results$counts,
    lda_varying_params_lists = lda_varying_params_lists,
    lda_fixed_params_list = list(method = "VEM"),
    dir = "../output/saved_models/runsFL/",
    reset = FALSE, #set this to FALSE after running once you don't need to do this everytime
    verbose = TRUE
  )

```


### for diagnostics
We will combine all the plots into one at the end of this section

align the topics based on different methods
```{r align-topics}

aligned_topics_product <- 
  align_topics(
    models = lda_models,
    method = "product") 

plot(aligned_topics_product, color_by = "coherence")

aligned_topics_transport <- 
  align_topics(
    models = lda_models,
    method = "transport") 

aligned_topics_transport@topics$coherence<-round(aligned_topics_transport@topics$coherence,1)
plot(aligned_topics_transport, color_by = "coherence",label_topics = TRUE)
```

### top taxa alingment - k10
```{r}
aligned_topics_transport_k <- 
  align_topics(
    models = lda_models[c(1,10)],
    method = "transport") 

p<-plot(aligned_topics_transport_k,
     add_leaves = TRUE, 
     label_topics = TRUE,
     n_features_in_leaves = 2,
     min_feature_prop = 0.01,
     leaves_text_size = 8,
     color_by="topic",
      model_name_repair_fun = function(x) str_remove_all(x, "k"))+
     theme_bw()+
    labs(x="")

#ggsave(p, filename = "fl_k10.png", height = 5, width = 9, units = "cm", scale = 1.5)

p
```



### perplexity
use to confirm that k=10 looks good. SEE THE bash script you should run on HPC as it takes too long on a personal computer.
https://stackoverflow.com/questions/21355156/topic-models-cross-validation-with-loglikelihood-or-perplexity
https://medium.com/@priyankads/perplexity-of-language-models-41160427ed72
https://datascience.stackexchange.com/questions/72882/topic-alignment-topic-modelling

Perplexity - This is a statistical measure of how well a probability model predicts a sample. As applied to LDA, for a given value of k, you estimate the LDA model. Then given the theoretical word distributions represented by the topics, compare that to the actual topic mixtures, or distribution of words in your documents. - Lower the better.

Result: We see the elbow at k=10 and this supports what we saw with ALTO so we proceed with K=10.

The lower the value the better the training we did! Look for the elbow in the plot.

```{r}
## read in perplexity result from your script 

FL_perplexity = read.csv("perplexity/FL_perplexity.csv",header=TRUE)

p<-ggplot(FL_perplexity, aes(x = k, y = perplexity)) +
   geom_point() +
   geom_smooth(se = FALSE) +
   theme_bw()+
   labs(x = "Candidate number of topics", y = "Perplexity score")

p
```

## diagnostic figure 
```{r align-topics-viz-figure-2, echo = FALSE, eval = FALSE}


make_figure2 <- function(aligned_topics_product,
                        aligned_topics_transport,perplexity_df,size_fraction){
  
  ## some edits
  aligned_topics_transport@topics$coherence<-round(aligned_topics_transport@topics$coherence,1)
  
  
  n_paths_combined <- 
    bind_rows(
      compute_number_of_paths(aligned_topics_product) %>% 
        mutate(method = "product"),
      compute_number_of_paths(aligned_topics_transport) %>% 
        mutate(method = "transport")
    )
  
  ## a
  g_key_topics <- 
    ggplot(n_paths_combined, 
           aes(x = n_topics,y = n_paths,col = method)) +
    geom_abline(slope = 1, intercept = 0, col = "gray90", size = 2) +
    geom_line() +
    geom_point(size = 1) +
    theme_minimal() +
    labs(x="# of topics in model",y="# of paths",title=size_fraction) +
    scale_x_continuous(breaks = 1:max(n_paths_combined$n_topics), minor_breaks = NULL) +
    scale_y_continuous(breaks = 1:max(n_paths_combined$n_topics), minor_breaks = NULL) +
    scale_color_manual("", values = c("coral","blue3")) + 
    theme_bw()+
    theme(
      legend.position= c(0.2, 0.8),
      legend.title = element_blank())
  
  
  topics <- 
    bind_rows(
      aligned_topics_product@topics %>% mutate(method = "product"),
      aligned_topics_transport@topics %>% mutate(method = "transport"),
    ) %>% 
    mutate(method = method %>% factor(., levels = c("transport","product"))) %>% 
    group_by(method, m) %>% mutate(K = n(), r = refinement) %>% ungroup() 
  
  model_summary <- 
    as.data.frame(topics) %>% 
   dplyr::group_by(method, K) %>%
    dplyr::summarize(
      minr = min(r),
      maxr = max(r), 
      r = median(r),
      .groups = "drop"
    )
  
  ##b
  g_coherence_transport <- 
    plot_alignment(
      aligned_topics_transport, 
      color_by = "co", 
      model_name_repair_fun = function(x) str_remove_all(x, "k"),
      label_topics = TRUE,
      n_features_in_leaves = 2
    ) + 
    theme_bw() +
    labs(y="coherence score\ntransport method", x= "# of topics in model")+
    theme(legend.position = "none")
  
  ##c
  g_refinement_distribution <- 
    ggplot(model_summary,
           aes(x = K, y = r)) + 
    geom_hline(yintercept = 0, col = "gray") +
    geom_ribbon(aes(ymin = minr, ymax = maxr), fill = "gray", alpha = 0.5) +
    geom_line(data = model_summary, col = "black", size = 1) +
    geom_point(data = model_summary, col = "black", size = 2) +
    geom_jitter(data = topics, aes(col = path), height = 0, width = 0.05, size = 1) +
    facet_wrap(method ~ ., dir = "v") + 
    scale_x_continuous(breaks = sort(unique(topics$K)), minor_breaks = NULL) +
    scale_y_log10() +
    scale_color_discrete(breaks = sort(unique(model_summary$K))) +
    guides(col = "none")  +
    xlab("# of topics in model") +
    ylab("refinement score") + theme_bw()+
    theme(axis.text.x = element_text(size=5))
  
  ##d
  perplexity<-ggplot(perplexity_df, aes(x = k, y = perplexity)) +
   geom_point() +
   geom_smooth(se = FALSE) +
   theme_bw()+
   labs(x = "cadidate # of topics", y = "Perplexity score")
  
  figure = 
    ggpubr::ggarrange(
      g_key_topics,
      g_coherence_transport,
      g_refinement_distribution,
      perplexity,
      ncol = 2, nrow = 2,
      heights = c(1, 1),
      widths = c(4,5),
      labels = c("a","b","c","d")
    )
  

  figure
}

figure <- make_figure2(aligned_topics_product = aligned_topics_product, 
                           aligned_topics_transport = aligned_topics_transport,
                       perplexity_df=FL_perplexity,size_fraction="Free-Living Prokaryotes")


ggsave(figure, filename = "output/plots/FL/diagnostics.png", height = 14, width = 21, units = "cm", scale = 1.5)

figure
```

panel a k=10-12 plateau 
panel b k=10-12 the scores are stable but break down (decraese after)
panel c transport plateau k=9-12, while the product keeps climbing not imformative
panel d perplexity elbow at a k=10 

k=10 was selected as the optimal number of topics based on convergent evidence from path stability, coherence, refinement, and perplexity diagnostics, with higher k values yielding fragmented, redundant topics.


### W and A
```{r}
W_FL = as.data.frame(lda_models$k10$gamma)
colnames(W_FL)<- sub("V", "", colnames(W_FL))
W_FL
```
```{r}
A <-as.data.frame(exp(lda_models$k10$beta))
A_df<-A ## to work on rare
A <- A %>% mutate(topic = row_number())

#change the ASV names 
A_long<- melt.data.frame(A, id.vars = "topic",variable_name="ASV")

# some cleaning up
#A_long$value<-round(A_long$value,5)
#A_long=subset(A_long, value>0)

#change the order column
A_long$Phyla<-sapply(strsplit(as.character(A_long$ASV)," "), "[", 1)
A_long$Order<-paste(sapply(strsplit(as.character(A_long$ASV)," "), "[", 2),
                    sapply(strsplit(as.character(A_long$ASV)," "), "[", 3),sep="_")
A_long$Order_raw<-paste(sapply(strsplit(as.character(A_long$Order),"_"), "[", 1))

#A_long_FL<- A_long
#save(A_long_FL,file="../output/tables/w_matrix/fl_tax")


A_long_FL <- A_long %>% mutate(asv_type=case_when(value >= 0.1 ~ ">=10%", value < 0.1 ~ "<10%"))
A_long_FL
```

#4. Run GoMDE
```{r}
library(fastTopics)
```

## fun:run_gomde_markers
```{r}
run_gomde_markers <- function(counts, tax_dict, W, A_long, lfsr_cut = 0.05, top_n = 30) {

  # align samples
  keep <- intersect(rownames(W), rownames(counts))
  W2 <- as.matrix(W[keep, , drop = FALSE])
  X2 <- as.matrix(counts[keep, , drop = FALSE])

  # normalize topic weights
  W2 <- W2 / rowSums(W2)

  # GoM DE
  de <- de_analysis(fit = W2, X = X2, pseudocount = 0.1, shrink.method = "ash",lfc.stat = "le",)
  #fits Poisson regressions for each ASV to estimate topic-specific parameters.
  #the default fit.method is scaled coordinate descent

  # tidy
  gomde <- as.data.frame(de$postmean) %>%
  rownames_to_column("ASV") %>%
  pivot_longer(-ASV, names_to = "topic", values_to = "log2FC") %>%
  left_join(
    as.data.frame(de$z) %>%
      rownames_to_column("ASV") %>%
      pivot_longer(-ASV, names_to = "topic", values_to = "z"),
    by = c("ASV","topic")
  ) %>%
  left_join(
    as.data.frame(de$lfsr) %>%
      rownames_to_column("ASV") %>%
      pivot_longer(-ASV, names_to = "topic", values_to = "lfsr"),
    by = c("ASV","topic")
  )


  #For each topic - we summarize the top_n most strongly enriched ASVs (by GoM DE effect size) as that topic’s diagnostic fingerprint.
  top_markers <- gomde %>%
  filter(lfsr < lfsr_cut, log2FC > 0) %>%
  group_by(topic) %>%
  slice_max(order_by = log2FC, n = top_n, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(tax_dict, by = c("ASV" = "ASV_name"))

  # overlap matrix
  marker_lists <- top_markers %>%
    group_by(topic) %>%
    summarise(asvs = list(unique(ASV))) %>%
    deframe()

  topic_names <- names(marker_lists)
  overlap_mat <- outer(
    topic_names, topic_names,
    Vectorize(function(t1, t2) length(intersect(marker_lists[[t1]], marker_lists[[t2]])))
  )
  rownames(overlap_mat) <- topic_names
  colnames(overlap_mat) <- topic_names

  list(
    de = de,
    gomde = gomde,
    top_markers = top_markers, ## gomde subsetted and + taxonomy 
    overlap_mat = overlap_mat ## print matrix of overlap
  )
}
```

```{r}
#res_FL  <- run_gomde_markers(FL_results$counts, FL_results$tax_dict,W_FL,A_long_FL,  top_n = 25)

#save so you don't have to re-computer
saveRDS(res_FL,"output/objects/FL/res_FL.rds")
res_FL<-readRDS("output/objects/FL/res_FL.rds")

res_FL
```
```{r}
subset(res_FL$gomde, lfsr <0.05 & log2FC>0 ) %>% 
  distinct(ASV) %>% 
  nrow()
```

```{r}
marker_phy_summary_FL <- res_FL$top_markers %>%
  count(topic, Phylum, sort = TRUE) %>%
  group_by(topic) %>%
  mutate(prop = n / sum(n)) %>%
  arrange(topic, desc(prop)) %>% 
  mutate(
    topic= factor(paste0("FL_", topic), levels = paste0("FL_", 1:10)),
    Phylum = ifelse(Phylum %in% names(phyla_colors), Phylum, "Other"),
    Phylum = factor(Phylum, levels =names(phyla_colors))
  )

marker_phy_summary_FL

```

## test to understand asvs in multiple topics
```{r}
pm_long<- as.data.frame(res_FL$de$postmean) %>%
  tibble::rownames_to_column("ASV") %>% 
  tidyr::pivot_longer(
    -ASV,
    names_to = "topic",
    values_to = "postmean"
  )

lfsr_long <- as.data.frame(res_FL$de$lfsr) %>%
  tibble::rownames_to_column("ASV") %>%
  tidyr::pivot_longer(-ASV, names_to = "topic", values_to = "lfsr")

pm_long <- left_join(pm_long, lfsr_long, by = c("ASV","topic"))

example_asv<-"Acidobacteriota (Bryobacterales) 1424"
# "Chloroflexi (Chloroflexales) 250"
#Acidobacteriota (Holophagales) 3011
# Acidobacteriota (Bryobacterales) 1424

pm_long %>%
  filter(ASV == example_asv) %>%
  arrange(desc(postmean))

pm_long

pm_long %>%
  filter(ASV == example_asv) 

pm_long %>%
  filter(ASV == example_asv) %>%
  ggplot(aes(topic, postmean)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = example_asv, y = "l.e. log2FC")
```


```{r}
get_le_comparisons <- function(de, asv, focal_topics = NULL) {
  if (is.null(de$F)) stop("de$F not found. Need fitted rates to reconstruct pairwise LFCs.")
  if (!asv %in% rownames(de$F)) stop("ASV not found in de$F rownames.")

  F_row <- de$F[asv, ]
  all_topics <- names(F_row)

  if (is.null(focal_topics)) focal_topics <- all_topics
  focal_topics <- intersect(focal_topics, all_topics)
  if (length(focal_topics) == 0) stop("No focal_topics matched topic names in de$F.")

  out <- lapply(focal_topics, function(j) {
    others <- setdiff(all_topics, j)
    lfc <- log2(F_row[j] / F_row[others])
    tibble(
      ASV = asv,
      focal = j,
      competitor = others,
      log2FC = as.numeric(lfc)
    ) %>%
      mutate(
        abs_log2FC = abs(log2FC),
        is_closest = abs_log2FC == min(abs_log2FC)
      ) %>%
      arrange(abs_log2FC)
  })

  bind_rows(out)
}

plot_le_comparisons <- function(df_comp,asv_name, n_show = NULL) {
  dfp <- df_comp
  if (!is.null(n_show)) {
    # keep the closest + the next (n_show-1) closest
    dfp <- dfp %>%
      group_by(focal) %>%
      slice_head(n = n_show) %>%
      ungroup()
  }

  dfp %>%
    group_by(focal) %>%
    mutate(competitor = factor(competitor, levels = competitor[order(log2FC)])) %>%
    ungroup() %>%
    ggplot(aes(x = competitor, y = log2FC, fill = is_closest)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_col() +
    coord_flip() +
    facet_wrap(~ focal, scales = "free_y") +
    theme_bw() +
    labs(
      x = "Competitor topic",
      y = "log2(F_focal / F_competitor)",
      fill = "Closest rival",
      title=paste0(asv_name)
    )
}
```

```{r}
example_asv <- "Acidobacteriota (Holophagales) 3011"
focals <- c("FL_6","FL_4", "FL_8")
#focals <- c("FL_7","FL_4")

df_comp <- get_le_comparisons(res_FL$de, example_asv, focal_topics = focals)

# table view: closest competitor per focal topic
df_comp %>%
  filter(is_closest) %>%
  select(ASV, focal, competitor, log2FC)

# plot all competitors (can be busy if many topics)
plot_le_comparisons(df_comp,example_asv)

```








## taxa distribution 
```{r}
res_FL$top_markers
```

```{r}
library(ComplexHeatmap)
library(circlize)
library(dplyr)

#grab data from full set not top25 -- but still only SPECIALIST
mat_df <- res_FL$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>% #sig enrichment 
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%      # specialists only
  ungroup() %>%
  left_join(FL_results$tax_dict, by = c("ASV" = "ASV_name")) %>%
  filter(!is.na(Order)) %>%
  filter(!Order %in% c("Incertae Sedis", "unclass_Bacteria")) %>%
  mutate(Phylum_clean = ifelse(Phylum %in% names(phyla_colors), Phylum, "Other")) %>%
  group_by(topic, Phylum, Phylum_clean, Order) %>%
  dplyr::summarise(n_asv = n(), .groups = "drop")


mat_df

## collapse orders we only see in 1 topic
min_asv <- 5  # orders with max < 3 across all topics become Other [Phylum]

# build matrix
# build full matrix first to know which orders are sparse
mat_full <- mat_df %>%
  group_by(Order, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order") %>%
  as.matrix()

# identify sparse orders
sparse_orders <- rownames(mat_full)[apply(mat_full, 1, max) < min_asv]

# collapse in mat_df
mat_df_collapsed <- mat_df %>%
  mutate(Order_plot = ifelse(Order %in% sparse_orders,
                           ifelse(Phylum_clean == "Other", 
                                  "Multiple (rare phyla)",  # catch the Other Other case this is when we made phyla other
                                  paste0("Multiple ", Phylum_clean)),
                           Order)) %>% 
  group_by(topic, Phylum_clean, Order_plot) %>%
  dplyr::summarise(n_asv = sum(n_asv), .groups = "drop")

# rebuild matrix
mat <- mat_df_collapsed %>%
  group_by(Order_plot, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order_plot") %>%
  as.matrix()

mat <- mat[, order(as.numeric(colnames(mat)))]
colnames(mat) <- paste0("FL_", colnames(mat))
mat <- mat[rowSums(mat) > 0, ]

# phylum annotation vector
# phylum lookup including Other [Phylum] rows
order_phylum <- mat_df_collapsed %>%
  distinct(Order_plot, Phylum_clean) %>%
  filter(Order_plot %in% rownames(mat))

phylum_vec <- setNames(order_phylum$Phylum_clean, order_phylum$Order_plot)
phylum_vec <- phylum_vec[rownames(mat)]

# lighten phylum colors for Other rows
lighten <- function(hex, factor = 0.5) {
  rgb <- col2rgb(hex)
  rgb_light <- rgb + (255 - rgb) * factor
  rgb(rgb_light[1,], rgb_light[2,], rgb_light[3,], maxColorValue = 255)
}

# build annotation colors - named orders get full color, Other gets lightened
row_colors <- sapply(rownames(mat), function(ord) {
  phylum <- phylum_vec[ord]
  base_col <- phyla_colors_clean[phylum]
  if (grepl("^Multiple ", ord)) lighten(base_col, 0.5) else base_col
})

# for ComplexHeatmap we need the annotation to use the phylum color vector
# but mark Other rows differently with a border or pattern - simplest is
# just to show them as a named color per row
row_annot <- rowAnnotation(
  Phylum = phylum_vec,
  col = list(Phylum = phyla_colors_clean),
  show_annotation_name = FALSE,
  annotation_legend_param = list(Phylum = list(ncol = 1)),
  show_legend = TRUE  
)


# color scale
col_fun <- colorRamp2(c(0, max(mat)), c("white", "#2C7FB8"))

row_title <- rowAnnotation(
  label = anno_text("Order", 
                    rot = 90, 
                    just = "center",
                    gp = gpar(fontsize = 10, fontface = "bold")),
  width = unit(0.8, "cm")
)

row_name_cols <- ifelse(grepl("^Multiple ", rownames(mat)), "grey50", "black")
row_name_fontface <- ifelse(grepl("^Multiple ", rownames(mat)), "italic", "plain")

ht_FL <- Heatmap(mat,
  name = "ASV count",
  col = col_fun,
  left_annotation = row_annot,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_dend = TRUE,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 8,
                      col = row_name_cols,
                      fontface = row_name_fontface),
  column_names_gp = gpar(fontsize = 10),
  column_title = expression(bold("A") * " Fingerprint Taxa"),
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  rect_gp = gpar(col = "grey90", lwd = 0.5),
  width = unit(10, "cm"),
  row_names_max_width = max_text_width(rownames(mat), gp = gpar(fontsize = 8)),
  layer_fun = function(j, i, x, y, width, height, fill) {
    v <- pindex(mat, i, j)
    grid.text(ifelse(v > 0, as.character(v), ""), x, y,
              gp = gpar(fontsize = 7, col = "grey20"))
  },
  row_title = "Order",
  row_title_gp = gpar(fontsize = 10, fontface = "bold"),
  row_title_side = "right"
) 

ggsave("output/plots/FL/GomDE_heat.pdf", fig4, width = 16, height = 8, units = "in", dpi = 600)

ht_FL

```

```{r}
row_annot2 <- rowAnnotation(
  Phylum = phylum_vec,
  col = list(Phylum = phyla_colors_clean),
  show_annotation_name = FALSE,  # no label
  show_legend = FALSE             # no legend
)

ht_FL2 <- Heatmap(mat,
  name = "ASV count",
  col = col_fun,
  left_annotation = row_annot2,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_dend = TRUE,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 8,
                      col = row_name_cols,
                      fontface = row_name_fontface),
  column_names_gp = gpar(fontsize = 10),
  column_title = expression(bold("A") * " Fingerprint Taxa"),
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  rect_gp = gpar(col = "grey90", lwd = 0.5),
  width = unit(10, "cm"),
  row_names_max_width = max_text_width(rownames(mat), gp = gpar(fontsize = 8)),
  layer_fun = function(j, i, x, y, width, height, fill) {
    v <- pindex(mat, i, j)
    grid.text(ifelse(v > 0, as.character(v), ""), x, y,
              gp = gpar(fontsize = 7, col = "grey20"))
  },
  row_title = "",
  row_title_gp = gpar(fontsize = 10, fontface = "bold"),
  row_title_side = "right"
)

ht_FL2
```


## top markers and all summary
```{r}
library(ggbeeswarm)
```


```{r}
df_swarm <- res_FL$top_markers %>%
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%        # pecialists only
  ungroup() %>%
  mutate(
    topic  = factor(topic),
    Phylum = ifelse(Phylum %in% names(phyla_colors), Phylum, "Other"),
    Phylum = factor(Phylum, levels = names(phyla_colors))
  )
## there are 2 --SEE HEATMAT BELOW

# pie composition per topic
pie_df <- df_swarm %>%
  count(topic, Phylum, name = "n") %>%
  group_by(topic) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  select(topic, Phylum, prop) %>%
  pivot_wider(names_from = Phylum, values_from = prop, values_fill = 0)

# CONTROL PIE LOCATION (global annotation band)
topic_levels <- levels(df_swarm$topic)
pie_df <- pie_df %>%
  mutate(x = as.numeric(factor(topic, levels = topic_levels)))

y_max <- max(df_swarm$log2FC, na.rm = TRUE)
pie_df$y <- y_max + 1.5   # <-- you can set this to whatever you want

pie_cols <- intersect(names(pie_df), names(phyla_colors))

## topic levels
topic_levels <- df_swarm %>%
  distinct(topic) %>%
  arrange(as.numeric(sub(".*_", "", topic))) %>%
  pull(topic)

df_swarm$topic <- factor(df_swarm$topic, levels = topic_levels)
pie_df$topic  <- factor(pie_df$topic,  levels = topic_levels)
pie_df$x <- as.numeric(pie_df$topic)


p_swarm_pie <- ggplot(df_swarm, aes(x = topic, y = log2FC, color = Phylum)) + #size = log2FC
  geom_quasirandom(width = 0.3, alpha = .8,stroke = 0.9 ) +
  scatterpie::geom_scatterpie(
    data = pie_df,
    aes(x = x, y = y),
    cols = pie_cols,
    pie_scale = 3,
    color = "white",
    linewidth = 0.3) +
  scale_color_manual(values = phyla_colors) +
  scale_fill_manual(values = phyla_colors) +
  guides(fill = "none", color = guide_legend(ncol = 4,byrow = TRUE)) + #size = guide_legend(ncol = 1)
  #guides(fill = "none")+
  theme_bw() +
  labs(x = "Topic", y = "GoM-DE log2FC - Top25 specilaists", color = "Phylum")+ #size = "log2FC"
  coord_fixed(ratio = 0.6, ylim = c(0, y_max + 2))+
  theme(
    panel.grid.major.x = element_blank(),
    legend.title = element_text(angle = 90),
    legend.spacing.x = unit(0.1, "mm"),
    legend.spacing.y = unit(0.1, "mm"),
     legend.key.size = unit(1, "mm"),
    legend.position = "bottom",
    legend.direction = "horizontal")

p_swarm_pie
```
## distribution of asvs heatmap
```{r}
asv_topic_sig <- res_FL$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>%            # enriched only
  left_join(distinct(FL_results$tax_dict, ASV_name, Phylum),
            by = c("ASV" = "ASV_name")) %>%
  mutate(
    Phylum = ifelse(Phylum %in% names(phyla_colors), Phylum, "Other"),
    Phylum = factor(Phylum, levels = names(phyla_colors))
  ) %>%
  group_by(ASV) %>%
  summarise(
    Phylum = dplyr::first(Phylum),
    n_topics = n_distinct(topic),
    max_log2FC = max(log2FC),
    .groups = "drop"
  )

df_hm <- asv_topic_sig %>%
  count(Phylum, n_topics)

pheat<-ggplot(df_hm, aes(x = factor(n_topics), y = Phylum, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  scale_fill_gradient(
    low = "#f2e5ff",
    high = "#5b2a86"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "# of topics where ASV is significantly enriched",
    y = "Phylum",
    fill = "ASV count",
     title = paste0(
      "Distribution of marker ASVs (n=",
      nrow(asv_topic_sig),
      ")"
     )
  )
pheat
```

```{r}
figure <-  ggpubr::ggarrange(pheat,p_swarm_pie, ncol = 2, nrow = 1,widths = c(.4,.6))

figure<-annotate_figure(figure,
                top = text_grob("Free-Living Prokaryotes", color = "black", face = "bold", size = 14))

ggsave(file="output/plots/FL/gomde_FL_pie.pdf", figure,width = 12, height = 8, units = "in",dpi=600)

figure
```

#5. Run Env Analysis

### topic proportion across samples
```{r}
paired_colors_num <- brewer.pal(10, "Paired")  # Get the colors
names(paired_colors_num) <- seq(1,10,1)
paired_colors_num
```

```{r}
## cluster by similarity
w_den<-as.dendrogram(as.hclust(agnes(W_FL, diss=F, method =  "ward")),hang=.5)

#data based on the clustering above 
W_FL2<-W_FL #Some we don't mess up the original
W_FL2$Sample_num<-row.names(W_FL)
W_long_FL<- melt.data.frame(W_FL2, id.vars="Sample_num", variable_name="topic")

plota_FL<-W_long_FL %>%
      ggplot(aes(x=factor(Sample_num,level=labels(w_den)), value, fill=topic))+
      geom_col(color=NA,width=1,linewidth=0)+
      scale_fill_manual(values=paired_colors_num)+
      labs(y="Subcommunity Proportion",
           x=paste0("Samples (n=", dim(W_FL2)[1],")"),fill="SC#",
           title=expression(bold("A") * ""))+ #Free Living Prokaryotes
      theme_bw()+
    scale_x_discrete(labels = NULL, breaks = NULL) + 
    scale_y_continuous(breaks = seq(0, 1, 0.1),expand = c(0, 0))+
      theme(axis.title.y = element_text(size = 10),
             plot.title = element_text(face = "bold", size = 10),
          axis.text.y = element_text(size = 10),
          legend.key.height= unit(.05, 'in'),
        legend.key.width= unit(.05, 'in'),
        legend.box.spacing = unit(0.01, "pt"))

ggsave("output/plots/FL/FL_structure.pdf", plota_FL, width = 7, height = 2.5, units = "in",dpi=600)
plota_FL

```

```{r}
FL_meta<-data.frame(FL_results$meta, check.names = FALSE)

#PLOTS B-D: legend bars
#get sample infor from main clean table
sample_info<-FL_meta[,c('Sample_num','new_depths_mdh2','CTD_TEMP','SEASON','YEAR','lake')]
#drop duplicates - because each OTU gets a value so we have many reps 
colnames(sample_info)<-c('sample_id','depth','CTD_TEMP','SEASON','YEAR','lake')
#reorder to match the order of the bar plots 
sample_info<-sample_info %>% arrange(sample_id,levels=labels(w_den))
sample_info$sample_id<- factor(sample_info$sample_id,level=labels(w_den))

###make the bar plots 
lake<-ggplot(data=sample_info, aes(x=sample_id, fill=lake)) + 
  scale_fill_manual(values=lake_colors)+
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

season<-ggplot(data=sample_info, aes(x=sample_id, fill=SEASON)) + 
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=season_colors)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

TEMP<-ggplot(data=sample_info, aes(x=sample_id, fill=CTD_TEMP)) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_gradient(low = "blue", high = "red")+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")
  
YEAR<-ggplot(data=sample_info, aes(x=sample_id, fill=factor(YEAR))) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=year_colors)+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

depth<-ggplot(data=sample_info, aes(x=sample_id, fill=depth)) + 
  scale_fill_manual(values=depth_colors_w)+
  geom_bar(width = 1,size=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

wx1=0.074
wx2=0.728

### value definitions
names_depth=c("spr:","I","E","T","H","smr:","E","D","T","H")

colors_depth= c(season_colors["Spring"],depth_colors_w["Iso_Top"], depth_colors_w["InvS_EPI"],
          depth_colors_w["InvS_TRM"], depth_colors_w["InvS_HYP"], season_colors["Summer"],
          depth_colors_w["Strat_EPI"],depth_colors_w["Strat_DCL"],depth_colors_w["Strat_TRM"],
          depth_colors_w["Strat_HYP"])

#change dynamically
label_start   <- wx1 + wx2 + 0.01
label_spacing <- 0.03

season_locx <- label_start + c(0, 2)                            * label_spacing  # 2 labels
lake_locx   <- label_start + c(0, 1, 2, 3, 4)                  * label_spacing  # 5 labels
strata_locx <- label_start + c(0, 0.03, 0.04, 0.05, 0.06, 0.09, 0.12, 0.13, 0.14, 0.15)
year_locx <- label_start + c(0, 0.02, 0.035, 0.05, 0.065, 0.08, 0.095, 0.11, 0.125) #8 labels
temp_locx   <- label_start + c(0, 1, 2)                        * label_spacing  # 3 labels


#gather all the plots
FL_structure<- ggdraw()+ theme_void()+
 draw_plot(plota_FL,x = 0, y = 0.25, width = 0.878, height = 0.75)+
 draw_plot(season,x = wx1, y= 0.2, width = wx2, height = 0.05)+
  draw_plot(lake,x = wx1, y=0.15, width = wx2, height = 0.05)+
  draw_plot(depth,x = wx1, y = 0.1, width =wx2, height = 0.05)+
  draw_plot(YEAR,x = wx1, y = 0.05, width =wx2, height = 0.05)+
  draw_plot(TEMP,x = wx1, y = 0, width =wx2, height = 0.05)+
  draw_plot_label(x = c(0.01,0.01,0.01,0.01,0.01), y = c(0.25,0.2,0.15,0.1,0.05), 
                     label =  c("Season","Lake","Depth","Year","Temp"),
                     hjust = 0, size = 7)+
  draw_plot_label(x=c(season_locx,lake_locx, strata_locx, year_locx, temp_locx), #temp
                  y=c(rep(0.25,2), #season x 2
                      rep(0.2,5), #lake x5
                      rep(0.15,10), #depth x 10
                      rep(0.1,9), #year x 9
                      rep(0.05,3)), #temp x3
                  label=c('Spring','Summer',
                          "SU","MI","HU","ER","ON",
                          names_depth,
                          "201","2","3","4","5","6","7","8","9",
                          'Low','High', 'NA'), 
                  color=c(season_colors["Spring"], season_colors["Summer"],
                          "#92809E", "#8cabea", "#64baaa","#FFC86A","#ee7061",
                          colors_depth,
                          "gray",'red','#A6CEE3','#B2DF8A',"brown","violet","purple","blue","black",
                          "blue","red","gray"),
                  hjust = 0, size = 6 )


  
  
ggsave("output/plots/FL/FL_sample_membership.pdf", FL_structure, width = 7, height = 3, units = "in",dpi=600)
  
FL_structure
```

### topic location
```{r}
#meta long
FL_meta_long<- melt.data.frame(FL_meta, id.vars="Sample_num", variable_name="variable")

highlighted_topics <- c(1,2,3,4,5,6,7,8,9,10)


W_long_meta_FL<-full_join(FL_meta_long, W_long_FL, by = "Sample_num") 

##to highlight a given topic we change alpha
W_long_meta_FL<- W_long_meta_FL %>% mutate(
    highlight = ifelse(as.character(topic) %in% highlighted_topics, "yes", "no")
  )


p_season<-subset(W_long_meta_FL,(variable == "SEASON") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(season_colors)),alpha = highlight))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=season_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Season",title="",y="Relative contribution",x= "Subcommunity")+
      theme(axis.title=element_text(size=8, face="bold"))

p_season

p_lake<-subset(W_long_meta_FL,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors)),alpha = highlight))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Lake",title="",y="",x= "")
p_lake

p_layer<-subset(W_long_meta_FL,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors)),alpha = highlight))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Layer",title="",y="",x= "")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )



p_layer

p_year<-subset(W_long_meta_FL,(variable == "YEAR") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(year_colors)),alpha = highlight))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=year_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Year",title="",y="",x= "")+
       theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )

p_year

p_status<-subset(W_long_meta_FL,(variable == "strat_status2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill= value.x,,alpha = highlight))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=strat_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      labs(x="",y="", fill="Strat. Type")+
      theme_classic()

p_status

# sample prop - how much does this sc represent overall 
hold<-subset(W_long_meta_FL, variable == "YEAR")  #can be any variable 
prop<-hold %>% group_by(topic) %>% dplyr::summarise(prop=round((sum(value.y)/sum(hold$value.y))*100))

p_dist<-ggplot(prop, aes(x=topic, y=prop, fill=topic,)) +
  geom_col() +
  geom_text(aes(label=paste0(round(prop, 1), "%")), 
            vjust=-0.5, 
            size=3, 
            fontface="bold") +  
  scale_fill_brewer(palette="Paired") +
  labs(x="", y="% of all sample distribution") +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))+
  theme(legend.position="none", 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(),
        theme(axis.title=element_text(size=8, face="bold")))


figure <- ggpubr::ggarrange(p_season, p_status, p_layer, p_lake, p_year,p_dist, ncol = 3, nrow = 2)
figure <-ggpubr::annotate_figure(figure, 
                        top = ggpubr::text_grob("FL PROK: Subcommunity distribution across samples", color = "black", 
                                        face = "bold", size = 12))

ggsave("output/plots/FL/FL_membership.pdf", figure, width = 9, height = 5, units = "in",dpi=600)

figure

```

```{r}

p_lake<-subset(W_long_meta_FL,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors)),alpha = highlight))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Lake",title=expression(bold("B") * " Meta Overview"), 
           y="Relative contribution",x= "Subcommunity")
p_lake

p_layer<-subset(W_long_meta_FL,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors)),alpha = highlight))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Layer",title="",y="Relative contribution",x= "Subcommunity")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )

FL_meta_plt<- ggpubr::ggarrange(p_lake, p_layer, ncol = 2, nrow = 1)
FL_meta_plt
```

### prep meta
```{r}
safe_logit <- function(x, eps = 1e-6) {
  x <- pmin(pmax(x, eps), 1 - eps)
  qlogis(x)
}

inv_logit <- function(x) {
  plogis(x)
}

keep=c("Alkalinity..Total.as.CaCO3_mg.l","Chlorophyll.a_ug.l",
       "Nitrogen..Total.Oxidized_mg.l","CTD_TEMP", "estimated_light_pcnt","Phosphorus..Total.as.P_ug.l",
      "Silica..Dissolved.as.Si_mg.l","Turbidity_NTU", "abs_depth_m", "STN_DEPTH_M", 
      "sbeox0Mg.L", "airT_14d_mean")

param_order <- c(
  "CTD_TEMP",
  "airT_14d_mean",
  "STN_DEPTH_M",
  "abs_depth_m",
  "sbeox0Mg.L",
  "estimated_light_logit",
  "Turbidity_NTU",
  "Chlorophyll.a_ug.l",
  "Alkalinity..Total.as.CaCO3_mg.l",
  "Nitrogen..Total.Oxidized_mg.l",
  "Phosphorus..Total.as.P_ug.l",
  "Silica..Dissolved.as.Si_mg.l")

meta_names<-read.csv("/Volumes/mlcoleman/project/greatlakes/Env_Metadata/meta_names.csv")

## we will filter samples out if meta is missing so we will only use complete cases 
prep_topic_meta <- function(W_X, X_meta,
                             na_thresh = 0.2,
                             light_var = "estimated_light_pcnt") {
  
  # check alignment
  if (!all(rownames(W_X) == rownames(X_meta))) {
    stop("rownames(W_X) and rownames(X_meta) are not identical and in the same order.")
  }
  
  # keep selected vars
  X_meta_num <- X_meta %>%
    dplyr::select(all_of(keep)) %>%
    dplyr::select(where(~ mean(is.na(.)) <= na_thresh))
  
  # transform light if present
  if (light_var %in% colnames(X_meta_num)) {
    X_meta_num <- X_meta_num %>%
      mutate(
        estimated_light_logit = safe_logit(.data[[light_var]] / 100)
      ) %>%
      select(-all_of(light_var))
  }
  
  # scale
  X_meta_num <- as.data.frame(scale(X_meta_num))
  rownames(X_meta_num) <- rownames(X_meta)
  
  # remove incomplete cases
  complete_rows <- complete.cases(X_meta_num)
  cat("Rows with NAs:", sum(!complete_rows), "\n")
  cat("Rows remaining:", sum(complete_rows), "\n")
  
  W_clean    <- W_X[complete_rows, ]
  meta_clean <- X_meta_num[complete_rows, ]
  
  list(
    W_clean      = W_clean,
    meta_clean   = meta_clean,
    meta_full    = X_meta_num,      # full metadata before removing NAs
    complete_rows = complete_rows   # logical index in case needed downstream
  )
}

prep_FL<- prep_topic_meta(W_FL, FL_meta)
prep_FL
```


### Limma
```{r}
#pak::pkg_install("bioc::limma")
library(limma)
```

Limma tests the observed effect. 
for each topic, does environment make that topic go up or down compared to all the other topics?

```{r}
run_topic_clr_limma <- function(W_X,
                                 X_meta_num,
                                 eps = 1e-6,
                                 add_intercept = TRUE) {

  # --- drop samples with missing metadata ---
  keep  <- complete.cases(X_meta_num)
  W_use <- W_X[keep, , drop = FALSE]
  X_use <- X_meta_num[keep, , drop = FALSE]

  # --- CLR transform across full composition ---
  W_clamped <- pmin(pmax(W_use, eps), 1 - eps)
  log_W     <- log(W_clamped)
  clr_W     <- log_W - rowMeans(log_W)   # subtract row-wise log geometric mean

  # limma wants features x samples
  Y_limma <- t(clr_W)

  # --- design matrix ---
  if (add_intercept) {
    design <- model.matrix(~ ., data = X_use)
  } else {
    design <- model.matrix(~ 0 + ., data = X_use)
  }

  # --- fit limma ---
  fit <- limma::lmFit(Y_limma, design)
  fit <- limma::eBayes(fit)

  # coefficients to report (drop intercept if present)
  coef_names <- colnames(design)
  if (add_intercept) {
    coef_names <- setdiff(coef_names, "(Intercept)")
  }

  # --- collect results per coefficient ---
  res_list <- lapply(coef_names, function(coef_nm) {
    tt <- limma::topTable(
      fit,
      coef   = coef_nm,
      number = Inf,
      sort.by = "none"
    )
    tt$topic <- rownames(tt)
    tt$term  <- coef_nm
    tt
  })

  res_df <- dplyr::bind_rows(res_list) %>%
    dplyr::as_tibble() %>%
    dplyr::rename(
      Estimate    = logFC,    # in CLR space: log-ratio effect vs geometric mean
      p_value     = P.Value,
      adj_p_value = adj.P.Val
    ) %>%
    dplyr::select(
      topic, term, Estimate, AveExpr, t, B,
      p_value, adj_p_value, everything()
    ) %>%
    dplyr::group_by(term) %>%
    dplyr::mutate(adj_p_value_by_term = p.adjust(p_value, method = "BH")) %>%
    dplyr::ungroup()

  list(
    clr_mat = clr_W,    # samples x topics in CLR space
    design  = design,
    fit     = fit,
    res_df  = res_df
  )
}

limma_res_FL_clr <- run_topic_clr_limma(prep_FL$W_clean, prep_FL$meta_clean)
sig_limma_clr_FL  <- limma_res_FL_clr$res_df  %>% filter(adj_p_value_by_term < 0.05)
sig_limma_clr_FL 
```


```{r}
plot_limma <- function(sig_df, topic_order = NULL, title = "Limma Results", 
                       meta_names = NULL, param_order = NULL) {
  
  # rename terms using meta_names if provided
  if (!is.null(meta_names)) {
    sig_df <- sig_df %>%
      mutate(term = trimws(term)) %>%
      left_join(meta_names %>% mutate(original = trimws(original)),
                by = c("term" = "original")) %>%
      mutate(term = coalesce(new, term))
  }
  
  # order topics
  if (!is.null(topic_order)) {
    sig_df <- sig_df %>%
      mutate(topic = factor(topic, levels = topic_order))
  }
  
  # translate param_order to new names if meta_names provided
  if (!is.null(param_order) && !is.null(meta_names)) {
    new_order <- meta_names$new[match(param_order, meta_names$original)]
    sig_df <- sig_df %>%
      mutate(term = factor(term, levels = new_order))
  } else if (!is.null(param_order)) {
    sig_df <- sig_df %>%
      mutate(term = factor(term, levels = param_order))
  } else {
    term_order <- sig_df %>%
      count(term) %>%
      arrange(n) %>%
      pull(term)
    sig_df <- sig_df %>%
      mutate(term = factor(term, levels = term_order))
  }
  
  ggplot(sig_df, aes(x = topic, y = term, fill = Estimate)) +
    geom_tile() +
    geom_text(aes(label = round(Estimate, 2)), size = 3) +
    scale_fill_gradient2(
      low      = "#3B82F6",
      mid      = "white",
      high     = "#EF4444",
      midpoint = 0
    ) +
    scale_y_discrete(drop = FALSE) +  # <-- keeps all factor levels even those without sig
    labs(
      x        = "Subcommunity",
      y        = "",
      fill     = "Estimate",
      title    = title,
      subtitle = "\n BH adj p < 0.05"
    ) +
    theme_bw() +
    theme(
      panel.grid      = element_blank(),
      axis.line       = element_line(colour = "black"),
      legend.position = "none", 
      plot.title      = element_text(face = "bold"),
    ) 
}

# call with meta_names and param_order passed in
FL_limma_crl <- plot_limma(
  sig_limma_clr_FL ,
  topic_order = colnames(prep_FL$W_clean),
  title       = expression(bold("C") * " Limma"),
  meta_names  = meta_names,
  param_order = param_order
)

FL_limma_crl

#ggsave("../output/plots/FL/FL_limma.png", FL_limma_crl, width = 8, height = 6, units = "in",dpi=600)
```



##RM + SHAP
We fit a Random Forest (RF) model to predict topic proportions from environmental metadata using the ranger package in R. Topic proportions were logit-transformed prior to fitting to address the bounded (0,1) nature of the response. For each topic, a separate RF model was trained with 500 trees, mtry = √p (where p is the number of predictors), and minimum node size of 5. Model performance was assessed using out-of-bag (OOB) R² (fit$r.squared from ranger), which provides a robust generalization estimate equivalent to LOO-CV — each sample is predicted only by trees for which it was not used in training (~37% of samples per tree, all samples predicted across the full forest). Variable importance was computed using impurity-based importance (mean decrease in node impurity). 
```{r}
library(ranger)
```
```{r}
run_rf_oob_clr <- function(W_clean, meta_clean) {
  
  # CLR transform on full composition matrix
  # log(x_i / geometric_mean(x)) for each document
  clr_transform <- function(W) {
    log_W     <- log(W + 1e-10)           # small epsilon guards any zeros
    gm        <- rowMeans(log_W)           # log geometric mean per document
    sweep(log_W, 1, gm, "-")              # subtract row-wise
  }
  
  clr_back_transform <- function(clr_W) {
    # softmax-style back transform to proportions
    exp_W <- exp(clr_W)
    sweep(exp_W, 1, rowSums(exp_W), "/")
  }
  
  W_clr     <- clr_transform(W_clean)
  topics    <- colnames(W_clean)
  n_samples <- nrow(W_clean)
  
  rf_pred   <- matrix(NA, nrow = n_samples, ncol = length(topics),
                      dimnames = list(rownames(W_clean), topics))
  rf_models <- list()
  
  for (topic in topics) {
    y  <- W_clr[, topic]                  # CLR-transformed outcome, unconstrained
    df <- data.frame(y = y, meta_clean)
    
    fit <- ranger::ranger(
      y ~ .,
      data          = df,
      num.trees     = 500,
      mtry          = floor(sqrt(ncol(meta_clean))),
      min.node.size = 5,
      importance    = "permutation",
      oob.error     = TRUE
    )
    
    rf_models[[topic]] <- fit
  }
  
  # collect CLR predictions then back-transform together
  # important: back-transform needs all topics simultaneously
  clr_pred <- W_clr
  for (topic in topics) {
    clr_pred[, topic] <- rf_models[[topic]]$predictions
  }
  rf_pred <- clr_back_transform(clr_pred)
  
  rf_perf <- data.frame(
    Topic = topics,
    R2    = sapply(rf_models, function(m) m$r.squared),
    RMSE  = sapply(topics, function(topic) {
      obs  <- W_clr[, topic]
      pred <- rf_models[[topic]]$predictions
      sqrt(mean((obs - pred)^2))
    })
  )
  rownames(rf_perf) <- NULL
  
  list(
    rf_pred   = rf_pred,       # back-transformed to proportion space
    clr_pred  = clr_pred,      # CLR space predictions, useful for diagnostics
    rf_models = rf_models,
    rf_perf   = rf_perf
  )
}

results_clr_FL <- run_rf_oob_clr(prep_FL$W_clean, prep_FL$meta_clean)
results_clr_FL
```

```{r}
plot_fit <- function(W_clean, pred_mat, perf_df, model_name = "RF") {
  
  topics <- colnames(W_clean)
  
  observed_long <- W_clean %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Sample") %>%
    tidyr::pivot_longer(-Sample, names_to = "Topic", values_to = "Observed")
  
  predicted_long <- pred_mat %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Sample") %>%
    tidyr::pivot_longer(-Sample, names_to = "Topic", values_to = "Predicted")
  
  plot_df <- left_join(observed_long, predicted_long, by = c("Sample", "Topic"))
  
  r2_labels <- perf_df %>%
    mutate(Topic = factor(Topic, levels = topics),
           label = paste0("R² = ", round(R2, 2)))
  
  ggplot(plot_df, aes(x = Observed, y = Predicted)) +
    geom_point(alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    facet_wrap(~ Topic, ncol = 5) +
    geom_text(
      data = r2_labels,
      aes(x = 0.5, y = Inf, label = label),
      hjust = 1.1, vjust = 1.2, inherit.aes = FALSE, size = 3
    ) +
    coord_equal() +
    theme_bw() +
    labs(
      x     = "Observed topic proportion",
      y     = paste(model_name, "predicted topic proportion"),
      title = paste(model_name)
    )
}

plot_fit(prep_FL$W_clean, results_clr_FL$rf_pred,  results_clr_FL$rf_perf,  model_name = "RF CLR")
```

```{r}
library(fastshap)
library(parallel)
```
shap - how much did this one variable contribute to pushing the prediction away from the average prediction?

SHapley Additive exPlanations

The theoretically exact SHAP value requires trying every possible combination of variables which is computationally explosive. fastshap uses a Monte Carlo approximation:
For each sample and each variable it:

Takes the real value of that variable for that sample
Randomly shuffles all other variables from other samples
Compares the prediction with the real value vs a random value for that variable
Repeats this nsim = 100 times and averages

The result is an estimate of "how much does knowing the true value of this variable change the prediction compared to not knowing it" — that's your SHAP value.

```{r}
run_shap_direction <- function(rf_models, meta_clean, nsim = 100, seed = 42, 
                                n_cores = parallel::detectCores() - 4) {
  
  pfun <- function(object, newdata) {
    ranger:::predict.ranger(object, data = newdata)$predictions
  }
  
  # Register parallel backend for fastshap
  cl <- parallel::makeCluster(n_cores)
  doParallel::registerDoParallel(cl)
  on.exit(parallel::stopCluster(cl))  # always cleans up, even on error
  
  purrr::map_dfr(names(rf_models), function(topic) {
    
    cat("Computing SHAP for", topic, "\n")
    
    set.seed(seed)
    shap_out <- fastshap::explain(
      object       = rf_models[[topic]],
      X            = meta_clean,
      pred_wrapper = pfun,
      nsim         = nsim,
      adjust       = TRUE,
      shap_only    = FALSE,
      parallel     = TRUE   # <-- now uses the registered cluster
    )
    
    shap_vals    <- as.data.frame(shap_out$shapley_values)
    feature_vals <- as.data.frame(shap_out$feature_values)
    
    purrr::map_dfr(colnames(shap_vals), function(env) {
      ct <- cor.test(feature_vals[[env]], shap_vals[[env]],
                     method = "spearman", use = "complete.obs")
      data.frame(
        Topic    = topic,
        Env      = env,
        shap_cor = as.numeric(ct$estimate),
        p_val    = ct$p.value,
        mean_abs = mean(abs(shap_vals[[env]]))
      )
    })
  }) %>%
    group_by(Topic) %>%
    mutate(
      p_adj      = p.adjust(p_val, method = "BH"),
      imp_rank   = rank(-mean_abs),
      imp_scaled = (mean_abs - min(mean_abs)) / (max(mean_abs) - min(mean_abs)),
      direction  = case_when(
        shap_cor >  0.1 ~ "+",
        shap_cor < -0.1 ~ "-",
        TRUE            ~ ""
      )
    ) %>%
    ungroup()
}
```

```{r}
# shap_dir_FL <- run_shap_direction(
#   rf_models  = results_clr_FL$rf_models,
#   meta_clean = prep_FL$meta_clean,
#   nsim       = 500
# )

#save so we don't need to re-run everytime
#saveRDS(shap_dir_FL, file = "output/objects/FL/shap_dir_FL.rds")
shap_dir_FL <- readRDS("output/objects/FL/shap_dir_FL.rds")

shap_dir_FL
```


```{r}
library(ggnewscale)
```
```{r}
plot_importance_direction_dot <- function(shap_dir, rf_perf,
                                          top_n = 5,
                                          p_cut = 1,
                                          h_line = 12.5,
                                          param_order,
                                          title = "SHAP Importance + Direction") {
  
  if (missing(param_order)) stop("param_order is required")
  
  topic_order <- unique(shap_dir$Topic)
  all_envs    <- param_order
  
  # ── identify top N env vars per topic by mean_abs SHAP ────────────────────
  top_vars <- shap_dir %>%
    group_by(Topic) %>%
    arrange(desc(mean_abs)) %>%
    slice(1:top_n) %>%
    ungroup() %>%
    select(Topic, Env) %>%
    mutate(is_top = TRUE)
  
  # ── build full grid ────────────────────────────────────────────────────────
  full_grid <- expand.grid(
    Topic = topic_order,
    Env   = all_envs,
    stringsAsFactors = FALSE
  )
  
  # ── join SHAP data + top_n flag onto full grid ─────────────────────────────
  plot_df <- full_grid %>%
    left_join(
      shap_dir %>% select(Topic, Env, imp_scaled, shap_cor, p_adj, mean_abs),
      by = c("Topic", "Env")
    ) %>%
    left_join(top_vars, by = c("Topic", "Env")) %>%
    mutate(
      # only show dot if in top_n AND significant
      imp_plot = ifelse(is.na(is_top) | p_adj >= p_cut, NA, imp_scaled),
      cor_plot = ifelse(is.na(is_top) | p_adj >= p_cut, NA, shap_cor),
      is_r2    = FALSE
    )
  
  # ── build RF R² row ────────────────────────────────────────────────────────
  r2_vals <- if ("RF_R2" %in% colnames(rf_perf)) {
    as.numeric(rf_perf$RF_R2)
  } else {
    as.numeric(rf_perf$R2)
  }
  
  r2_row <- data.frame(
    Topic    = rf_perf$Topic,
    Env      = "RF R²",
    imp_plot = r2_vals,
    cor_plot = NA,
    is_r2    = TRUE
  )
  
  # ── combine ────────────────────────────────────────────────────────────────
  combined_df <- bind_rows(plot_df, r2_row) %>%
    mutate(
      Env   = factor(Env,   levels = c("RF R²", all_envs)),
      Topic = factor(Topic, levels = topic_order)
    )
  
  imp_df_plot <- combined_df %>% filter(!is_r2)
  r2_df_plot  <- combined_df %>% filter(is_r2)
  
  ggplot() +
    # ── background tiles ──────────────────────────────────────────────────
    geom_tile(data  = imp_df_plot,
              aes(x = Topic, y = Env),
              fill  = "white", color = "grey90") +
    
    # ── dots: size = importance, color = shap_cor ─────────────────────────
    geom_point(data = imp_df_plot %>% filter(!is.na(imp_plot)),
               aes(x     = Topic, y    = Env,
                   size  = imp_plot,
                   color = cor_plot)) +
    scale_color_gradient2(
      low      = "#2C7BB6",
      mid      = "white",
      high     = "#D7191C",
      midpoint = 0,
      name     = "Spearman ρ\n(feature × SHAP)",
      limits   = c(-1, 1)
    ) +
    scale_size_continuous(
      name   = "Scaled SHAP\nimportance",
      range  = c(2, 8),
      limits = c(0, 1)
    ) +
    
    # ── R² row ────────────────────────────────────────────────────────────
    new_scale_fill() +
    geom_tile(data  = r2_df_plot,
              aes(x = Topic, y = Env, fill = imp_plot),
              color = "white") +
    scale_fill_gradient(low    = "#EFF3FF", high   = "darkgreen",
                        name   = "RF OOB R²",
                        limits = c(0, 1)) +
    geom_text(data = r2_df_plot,
              aes(x = Topic, y = Env,
                  label = round(imp_plot, 2)),
              size = 4, color = "black") +
    
    geom_hline(yintercept = h_line, color = "black", linewidth = 0.8) +
    
    labs(
      x        = "Subcommunity",
      y        = "",
      title    = title,
      #subtitle = paste0("Top ", top_n, " drivers per topic")
    ) +
    theme_bw() +
    theme(
      panel.grid   = element_blank(),
      axis.line    = element_line(colour = "black"),
      plot.title   = element_text(face = "bold", size = 12),
      legend.box   = "vertical"
    ) +
    guides(fill = "none")
}

# get the new names in the same order as param_order
new_order <- meta_names$new[match(param_order, meta_names$original)]

shap_dir_FL2<- shap_dir_FL %>%
  left_join(meta_names, by = c("Env" = "original")) %>%
  mutate(Env= coalesce(new, Env)) %>% 
  mutate(Env = factor(Env, levels = new_order))

FL_rfshap<-plot_importance_direction_dot(
  shap_dir    = shap_dir_FL2,
  rf_perf     = results_clr_FL$rf_perf,
  top_n       = 15,
  param_order = new_order,
  title       = expression(bold("D") * " RM + SHAP"), 
)

ggsave("../output/plots/FL/FL_RFshap.png", FL_rfshap, width = 8, height = 6, units = "in",dpi=600)

FL_rfshap
```
Direction was summarized as the Spearman correlation between each environmental variable and its SHapley Additive exPlanation (SHAP) value, where positive values indicate the variable drives topic abundance upward.

#Summary plot
```{r}
FL_limma_crl_tight <- FL_limma_crl + theme(plot.margin = margin(0, -2, 0, 0))
FL_rfshap_tight    <- FL_rfshap    + theme(plot.margin = margin(0, 0, 0, -2))
bottom_right <- plot_grid(FL_limma_crl_tight, FL_rfshap_tight, ncol = 2,rel_widths = c(.8, 1.2))


right_col <- plot_grid(FL_meta_plt, bottom_right, nrow = 2,rel_heights = c(.8, 1.3))

#grab ht_FL as a grob if it's a ComplexHeatmap
ht_grob <- grid::grid.grabExpr(ComplexHeatmap::draw(ht_FL))
FL_summary <- plot_grid(ht_grob, right_col, 
                        ncol = 2,
                        align = "h",
                        axis = "tb",
                        rel_widths = c(1, 1.3))

ggsave("output/plots/FL/FL_summary.pdf", FL_summary, width = 18, height = 8, units = "in",dpi=600)
FL_summary
```


#____________
#**PA**
#2. Prep data 
```{r}
#this inlcudes counts, tax_dict, meta
PA_results<-process_phyloseq(PA_data)

#save the counts tables you will need for perplexity
write.csv(PA_results$counts, "output/tables/runPA/PA_counts.csv")

#free up some memoery we don't need the FL_data anymore and we can call it in if we do 
rm(PA_data)
gc()  

```

#3. Run LDA
### run LDA models 
```{r run-lda-models}
lda_varying_params_lists_PA <-  list()
for (k in 1:20) {lda_varying_params_lists_PA[[paste0("k",k)]] <- list(k = k)}

lda_models_PA  <- 
  run_lda_models(
    data = PA_results$counts,
    lda_varying_params_lists = lda_varying_params_lists_PA,
    lda_fixed_params_list = list(method = "VEM"),
    dir = "../output/saved_models/runsPA/",
    reset = FALSE,
    verbose = TRUE
  )

```

### align the topics based on different methods
```{r align-topics}

aligned_topics_product_PA <- 
  align_topics(
    models = lda_models_PA,
    method = "product") 

compute_number_of_paths(aligned_topics_product_PA) %>% 
  plot_number_of_paths() + 
  ggtitle("Method: product")

aligned_topics_transport_PA <- 
  align_topics(
    models = lda_models_PA,
    method = "transport") 

compute_number_of_paths(aligned_topics_transport_PA) %>% 
  plot_number_of_paths() + 
  ggtitle("Method: transport")

```

### top taxa alingment
```{r}
aligned_topics_transport_k <- 
  align_topics(
    models = lda_models_PA[c(1,10)],
    method = "transport") 

p<-plot(aligned_topics_transport_k,
     add_leaves = TRUE, 
     label_topics = TRUE,
     n_features_in_leaves = 2,
     min_feature_prop = 0.01,
     leaves_text_size = 8,
     color_by="topic",
      model_name_repair_fun = function(x) str_remove_all(x, "k"))+
     theme_bw()+
    labs(x="")

#ggsave(p, filename = "PA_k10.png", height = 5, width = 9, units = "cm", scale = 1.5)

p
```

### perplexity (for diagnostics)
```{r}
PA_perplexity = read.csv("perplexity/PA_perplexity.csv",header=TRUE)

p<-ggplot(PA_perplexity, aes(x = k, y = perplexity)) +
   geom_point() +
   geom_smooth(se = FALSE) +
   theme_bw()+
   labs(x = "Candidate number of topics", y = "Perplexity score")

p
```

##diagnostics figure
```{r}
figure <- make_figure2(aligned_topics_product = aligned_topics_product_PA, 
                           aligned_topics_transport = aligned_topics_transport_PA,
                       perplexity_df=PA_perplexity,size_fraction="Particle-Associated Prokaryotes")


ggsave(figure, filename = "output/plots/PA/diagnostics.png", height = 14, width = 21, units = "cm", scale = 1.5)

figure

```

### W and A
```{r}
W_PA = as.data.frame(lda_models_PA$k10$gamma)
colnames(W_PA)<- sub("V", "", colnames(W_PA))
W_PA
```

```{r}
A_PA <-as.data.frame(exp(lda_models_PA$k10$beta))
A_df_PA<-A_PA ## to work on rare
A_PA <- A_PA %>% mutate(topic = row_number())

#change the ASV names 
A_long_PA<- melt.data.frame(A_PA, id.vars = "topic",variable_name="ASV")
A_long_PA$Phyla<-sapply(strsplit(as.character(A_long_PA$ASV)," "), "[", 1)
A_long_PA$Order<-paste(sapply(strsplit(as.character(A_long_PA$ASV)," "), "[", 2),
                    sapply(strsplit(as.character(A_long_PA$ASV)," "), "[", 3),sep="_")
A_long_PA$Order_raw<-paste(sapply(strsplit(as.character(A_long_PA$Order),"_"), "[", 1))

#A_long_PA<- A_long_PA
#save(A_long_PA,file="../output/tables/w_matrix/pa_tax")


A_long_PA <- A_long_PA %>% mutate(asv_type=case_when(value >= 0.1 ~ ">=10%", value < 0.1 ~ "<10%"))
A_long_PA
```


#4.Run GoMDE
```{r}
#res_PA  <- run_gomde_markers(PA_results$counts, PA_results$tax_dict,W_PA,A_long_PA,  top_n = 25)

#save so you don't have to re-computer
saveRDS(res_PA,"output/objects/PA/res_PA.rds")
res_PA<-readRDS("output/objects/PA/res_PA.rds")

res_PA
```

### overview
```{r}
#grab data from full set not top25 -- but still only SPECIALIST
mat_df <- res_PA$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>% #sig enrichment 
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%      # specialists only
  ungroup() %>%
  left_join(PA_results$tax_dict, by = c("ASV" = "ASV_name")) %>%
  filter(!is.na(Order)) %>%
  filter(!Order %in% c("Incertae Sedis", "unclass_Bacteria")) %>%
  mutate(Phylum_clean = ifelse(Phylum %in% names(phyla_colors), Phylum, "Other")) %>%
  group_by(topic, Phylum, Phylum_clean, Order) %>%
  dplyr::summarise(n_asv = n(), .groups = "drop")


mat_df

## collapse orders we only see in 1 topic
min_asv <- 5  # orders with max < 3 across all topics become Other [Phylum]

# build matrix
# build full matrix first to know which orders are sparse
mat_full <- mat_df %>%
  group_by(Order, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order") %>%
  as.matrix()

# identify sparse orders
sparse_orders <- rownames(mat_full)[apply(mat_full, 1, max) < min_asv]

# collapse in mat_df
mat_df_collapsed <- mat_df %>%
  mutate(Order_plot = ifelse(Order %in% sparse_orders,
                           ifelse(Phylum_clean == "Other", 
                                  "Other (rare phyla)",  # catch the Other Other case this is when we made phyla other
                                  paste0("Multiple ", Phylum_clean)),
                           Order)) %>% 
  group_by(topic, Phylum_clean, Order_plot) %>%
  dplyr::summarise(n_asv = sum(n_asv), .groups = "drop")

# rebuild matrix
mat <- mat_df_collapsed %>%
  group_by(Order_plot, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order_plot") %>%
  as.matrix()

mat <- mat[, order(as.numeric(colnames(mat)))]
colnames(mat) <- paste0("PA_", colnames(mat))
mat <- mat[rowSums(mat) > 0, ]

# phylum annotation vector
# phylum lookup including Other [Phylum] rows
order_phylum <- mat_df_collapsed %>%
  distinct(Order_plot, Phylum_clean) %>%
  filter(Order_plot %in% rownames(mat))

phylum_vec <- setNames(order_phylum$Phylum_clean, order_phylum$Order_plot)
phylum_vec <- phylum_vec[rownames(mat)]

# lighten phylum colors for Other rows
lighten <- function(hex, factor = 0.5) {
  rgb <- col2rgb(hex)
  rgb_light <- rgb + (255 - rgb) * factor
  rgb(rgb_light[1,], rgb_light[2,], rgb_light[3,], maxColorValue = 255)
}

# build annotation colors - named orders get full color, Other gets lightened
row_colors <- sapply(rownames(mat), function(ord) {
  phylum <- phylum_vec[ord]
  base_col <- phyla_colors_clean[phylum]
  if (grepl("^Multiple ", ord)) lighten(base_col, 0.5) else base_col
})

# for ComplexHeatmap we need the annotation to use the phylum color vector
# but mark Other rows differently with a border or pattern - simplest is
# just to show them as a named color per row
row_annot <- rowAnnotation(
  Phylum = phylum_vec,
  col = list(Phylum = phyla_colors_clean),
  show_annotation_name = FALSE,
  annotation_legend_param = list(Phylum = list(ncol = 1)),
  show_legend = TRUE  
)


# color scale
col_fun <- colorRamp2(c(0, max(mat)), c("white", "#2C7FB8"))

row_title <- rowAnnotation(
  label = anno_text("Order", 
                    rot = 90, 
                    just = "center",
                    gp = gpar(fontsize = 10, fontface = "bold")),
  width = unit(0.8, "cm")
)

row_name_cols <- ifelse(grepl("^Multiple", rownames(mat)), "grey50", "black")
row_name_fontface <- ifelse(grepl("^Multiple ", rownames(mat)), "italic", "plain")

ht_PA <- Heatmap(mat,
  name = "ASV count",
  col = col_fun,
  left_annotation = row_annot,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_dend = TRUE,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 8,
                      col = row_name_cols,
                      fontface = row_name_fontface),
  column_names_gp = gpar(fontsize = 10),
  column_title = expression(bold("A") * " Fingerprint Taxa"),
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  rect_gp = gpar(col = "grey90", lwd = 0.5),
  width = unit(10, "cm"),
  row_names_max_width = max_text_width(rownames(mat), gp = gpar(fontsize = 8)),
  layer_fun = function(j, i, x, y, width, height, fill) {
    v <- pindex(mat, i, j)
    grid.text(ifelse(v > 0, as.character(v), ""), x, y,
              gp = gpar(fontsize = 7, col = "grey20"))
  },
  row_title = "Order",
  row_title_gp = gpar(fontsize = 10, fontface = "bold"),
  row_title_side = "right"
) 
  
ht_PA

```

### top markers summary
```{r}
df_swarm <- res_PA$top_markers %>%
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%        # pecialists only
  ungroup() %>%
  mutate(
    topic  = factor(topic),
    Phylum = ifelse(Phylum %in% names(phyla_colors), Phylum, "Other"),
    Phylum = factor(Phylum, levels = names(phyla_colors))
  )

# pie composition per topic
pie_df <- df_swarm %>%
  count(topic, Phylum, name = "n") %>%
  group_by(topic) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  select(topic, Phylum, prop) %>%
  pivot_wider(names_from = Phylum, values_from = prop, values_fill = 0)

# CONTROL PIE LOCATION (global annotation band)
topic_levels <- levels(df_swarm$topic)
pie_df <- pie_df %>%
  mutate(x = as.numeric(factor(topic, levels = topic_levels)))

y_max <- max(df_swarm$log2FC, na.rm = TRUE)
pie_df$y <- y_max + 1.5   # <-- you can set this to whatever you want

pie_cols <- intersect(names(pie_df), names(phyla_colors))

## topic levels
topic_levels <- df_swarm %>%
  distinct(topic) %>%
  arrange(as.numeric(sub(".*_", "", topic))) %>%
  pull(topic)

df_swarm$topic <- factor(df_swarm$topic, levels = topic_levels)
pie_df$topic  <- factor(pie_df$topic,  levels = topic_levels)
pie_df$x <- as.numeric(pie_df$topic)


p_swarm_pie_pa <- ggplot(df_swarm, aes(x = topic, y = log2FC, color = Phylum)) + #size = log2FC
  geom_quasirandom(width = 0.3, alpha = .8,stroke = 0.9 ) +
  scatterpie::geom_scatterpie(
    data = pie_df,
    aes(x = x, y = y),
    cols = pie_cols,
    pie_scale = 3,
    color = "white",
    linewidth = 0.3) +
  scale_color_manual(values = phyla_colors) +
  scale_fill_manual(values = phyla_colors) +
  guides(fill = "none", color = guide_legend(ncol = 4,byrow = TRUE)) + #size = guide_legend(ncol = 1)
  #guides(fill = "none")+
  theme_bw() +
  labs(x = "Topic", y = "GoM-DE log2FC - Top25 Specialists", color = "Phylum")+ #size = "log2FC"
  coord_fixed(ratio = 0.6, ylim = c(0, y_max + 2))+
  theme(
    panel.grid.major.x = element_blank(),
    legend.title = element_text(angle = 90),
    legend.spacing.x = unit(0.1, "mm"),
    legend.spacing.y = unit(0.1, "mm"),
     legend.key.size = unit(1, "mm"),
    legend.position = "bottom",
    legend.direction = "horizontal")

p_swarm_pie_pa
```
```{r}
asv_topic_sig_pa <- res_PA$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>%            # enriched only
  left_join(distinct(PA_results$tax_dict, ASV_name, Phylum),
            by = c("ASV" = "ASV_name")) %>%
  mutate(
    Phylum = ifelse(Phylum %in% names(phyla_colors), Phylum, "Other"),
    Phylum = factor(Phylum, levels = names(phyla_colors))
  ) %>%
  group_by(ASV) %>%
  summarise(
    Phylum = dplyr::first(Phylum),
    n_topics = n_distinct(topic),
    max_log2FC = max(log2FC),
    .groups = "drop"
  )

df_hm_pa <- asv_topic_sig_pa %>%
  count(Phylum, n_topics)
print(sum(df_hm_pa$n))

pheat_pa<-ggplot(df_hm_pa, aes(x = factor(n_topics), y = Phylum, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  scale_fill_gradient(
    low = "#f2e5ff",
    high = "#5b2a86"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Number of topics where ASV is significantly enriched",
    y = "Phylum",
    fill = "ASV count",
     title = paste0(
      "Distribution of marker ASVs \n (n=",
      nrow(asv_topic_sig_pa),
      ")"
     )
  )
pheat_pa
```
```{r}
figure <-  ggpubr::ggarrange(pheat_pa,p_swarm_pie_pa, ncol = 2, nrow = 1,widths = c(.4,.6))

figure<-annotate_figure(figure,
                top = text_grob("Particle-Associated Prokaryotes", color = "black", face = "bold", size = 14))

ggsave(file="output/plots/PA/gomde_PA_pie.pdf", figure, width = 12, height = 8, units = "in",dpi=600)

figure
```



#5.Run Env Analysis

### topic proportion across samples
```{r}
## cluster by similarity
w_den<-as.dendrogram(as.hclust(agnes(W_PA, diss=F, method =  "ward")),hang=.5)

#data based on the clustering above 
W_PA2<-W_PA
W_PA2$Sample_num<-row.names(W_PA2)
W_long_PA<- melt.data.frame( W_PA2, id.vars="Sample_num", variable_name="topic")

plota_PA<-W_long_PA %>%
      ggplot(aes(x=factor(Sample_num,level=labels(w_den)), value, fill=topic))+
      geom_col(color=NA,width=1,linewidth=0)+
      scale_fill_brewer(palette="Paired")+
      labs(y="Subcommunity Proportion",
           x=paste0("Samples (n=", dim(W_PA2)[1],")"),fill="SC#",
           title=expression(bold("B") * " PA Prokaryotes"))+
      theme_bw()+
    scale_x_discrete(labels = NULL, breaks = NULL) + 
    scale_y_continuous(breaks = seq(0, 1, 0.1),expand = c(0, 0))+
      theme(axis.title.y = element_text(size = 10),
             plot.title = element_text(face = "bold", size = 12),
          axis.text.y = element_text(size = 10),
          legend.key.height= unit(.05, 'in'),
        legend.key.width= unit(.05, 'in'),
        legend.box.spacing = unit(0.01, "pt"))

plota_PA

ggsave("output/plots/PA/PA_structure.pdf", plota_PA, width = 7, height = 2.5, units = "in",dpi=600)

```

```{r}
PA_meta<-data.frame(PA_results$meta, check.names = FALSE)

#PLOTS B-D: legend bars
#get sample infor from main clean table
sample_info<-PA_meta[,c('Sample_num','new_depths_mdh2','CTD_TEMP','SEASON','YEAR','lake')]
#drop duplicates - because each OTU gets a value so we have many reps 
colnames(sample_info)<-c('sample_id','depth','CTD_TEMP','SEASON','YEAR','lake')
#reorder to match the order of the bar plots 
sample_info<-sample_info %>% arrange(sample_id,levels=labels(w_den))
sample_info$sample_id<- factor(sample_info$sample_id,level=labels(w_den))

###make the bar plots 
lake<-ggplot(data=sample_info, aes(x=sample_id, fill=lake)) + 
  scale_fill_manual(values=lake_colors)+
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

season<-ggplot(data=sample_info, aes(x=sample_id, fill=SEASON)) + 
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=season_colors)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

TEMP<-ggplot(data=sample_info, aes(x=sample_id, fill=CTD_TEMP)) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_gradient(low = "blue", high = "red")+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")
  
YEAR<-ggplot(data=sample_info, aes(x=sample_id, fill=factor(YEAR))) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=year_colors)+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

depth<-ggplot(data=sample_info, aes(x=sample_id, fill=depth)) + 
  scale_fill_manual(values=depth_colors_w)+
  geom_bar(width = 1,size=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

wx1=0.072
wx2=0.73

### value definitions
names_depth=c("spr:","I","E","T","H","smr:","E","D","T","H")

colors_depth= c(season_colors["Spring"],depth_colors_w["Iso_Top"], depth_colors_w["InvS_EPI"],
          depth_colors_w["InvS_TRM"], depth_colors_w["InvS_HYP"], season_colors["Summer"],
          depth_colors_w["Strat_EPI"],depth_colors_w["Strat_DCL"],depth_colors_w["Strat_TRM"],
          depth_colors_w["Strat_HYP"])

#change dynamically
label_start   <- wx1 + wx2 + 0.01
label_spacing <- 0.03

season_locx <- label_start + c(0, 2)                            * label_spacing  # 2 labels
lake_locx   <- label_start + c(0, 1, 2, 3, 4)                  * label_spacing  # 5 labels
strata_locx <- label_start + c(0, 0.03, 0.04, 0.05, 0.06, 0.09, 0.12, 0.13, 0.14, 0.15)
year_locx <- label_start + c(0, 0.02, 0.035, 0.05, 0.065, 0.08, 0.095, 0.11, 0.125) #8 labels
temp_locx   <- label_start + c(0, 1, 2)                        * label_spacing  # 3 labels


#gather all the plots
PA_structure<- ggdraw()+ theme_void()+
 draw_plot(plota_PA,x = 0, y = 0.25, width = 0.878, height = 0.75)+
 draw_plot(season,x = wx1, y= 0.2, width = wx2, height = 0.05)+
  draw_plot(lake,x = wx1, y=0.15, width = wx2, height = 0.05)+
  draw_plot(depth,x = wx1, y = 0.1, width =wx2, height = 0.05)+
  draw_plot(YEAR,x = wx1, y = 0.05, width =wx2, height = 0.05)+
  draw_plot(TEMP,x = wx1, y = 0, width =wx2, height = 0.05)+
  draw_plot_label(x = c(0.01,0.01,0.01,0.01,0.01), y = c(0.25,0.2,0.15,0.1,0.05), 
                     label =  c("Season","Lake","Depth","Year","Temp"),
                     hjust = 0, size = 7)+
  draw_plot_label(x=c(season_locx,lake_locx, strata_locx, year_locx, temp_locx), #temp
                  y=c(rep(0.25,2), #season x 2
                      rep(0.2,5), #lake x5
                      rep(0.15,10), #depth x 10
                      rep(0.1,9), #year x 9
                      rep(0.05,3)), #temp x3
                  label=c('Spring','Summer',
                          "SU","MI","HU","ER","ON",
                          names_depth,
                          "201","2","3","4","5","6","7","8","9",
                          'Low','High', 'NA'), 
                  color=c(season_colors["Spring"], season_colors["Summer"],
                          "#92809E", "#8cabea", "#64baaa","#FFC86A","#ee7061",
                          colors_depth,
                          "gray",'red','#A6CEE3','#B2DF8A',"brown","violet","purple","blue","black",
                          "blue","red","gray"),
                  hjust = 0, size = 6 )


  
  
ggsave("output/plots/PA/PA_sample_membership.pdf", PA_structure, width = 7, height = 3, units = "in",dpi=600)
  
PA_structure
```




### topic location
```{r}
#meta long
PA_meta_long<- melt.data.frame(subset(PA_meta,strat_status != "missing_data"), id.vars="Sample_num", variable_name="variable")
W_long_meta_PA<-full_join(PA_meta_long, W_long_PA, by = "Sample_num")

p_season<-subset(W_long_meta_PA,(variable == "SEASON") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(season_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=season_colors)+
      theme_bw()+
      labs(fill="Season",title="",y="Relative contribution",x= "Subcommunity")

p_season

p_lake<-subset(W_long_meta_PA,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      theme_bw()+
      labs(fill="Lake",title="",y="",x= "")
p_lake

p_layer<-subset(W_long_meta_PA,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      theme_classic()+
      labs(fill="Layer",title="",y="",x= "")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )



p_layer

p_year<-subset(W_long_meta_PA,(variable == "YEAR") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(year_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=year_colors)+
      theme_bw()+
      labs(fill="Year",title="",y="",x= "")

p_year

p_status<-subset(W_long_meta_PA,(variable == "strat_status2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill= value.x))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=strat_colors)+
      labs(x="",y="", fill="Strat. Type")+
      theme_bw()

p_status

#sample prop - how much does this sc represent overall 
hold<-subset(W_long_meta_PA, variable == "YEAR")  #can be any variable 
prop<-hold %>% group_by(topic) %>% dplyr::summarise(prop=round((sum(value.y)/sum(hold$value.y))*100))

p_dist<-ggplot(prop, aes(x=topic, y=prop, fill=topic)) +
  geom_col() +
  geom_text(aes(label=paste0(round(prop, 1), "%")), 
            vjust=-0.5, 
            size=3, 
            fontface="bold") +  
  scale_fill_brewer(palette="Paired") +
  labs(x="", y="% of all sample distribution") +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))+
  theme(legend.position="none", 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(),
        theme(axis.title=element_text(size=8, face="bold")))


figure = ggpubr::ggarrange(p_season, p_status, p_layer, p_lake, p_year,p_dist, ncol = 3, nrow = 2)

figure<-ggpubr::annotate_figure(figure, 
                        top = ggpubr::text_grob("PA PROK: Topic Distribution across samples", color = "black", 
                                        face = "bold", size = 12))

ggsave("output/plots/PA/PA_membership.pdf", figure, width = 9, height = 5, units = "in",dpi=600)

figure

```


```{r}

p_lake<-subset(W_long_meta_PA,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Lake",title=expression(bold("B") * " Meta Overview"), 
           y="Relative contribution",x= "Subcommunity")
p_lake

p_layer<-subset(W_long_meta_PA,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Layer",title="",y="Relative contribution",x= "Subcommunity")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )

PA_meta_plt<- ggpubr::ggarrange(p_lake, p_layer, ncol = 2, nrow = 1)
PA_meta_plt
```

##prep meta
```{r}
prep_PA   <- prep_topic_meta(W_PA, PA_meta)
```

## Limma
```{r}
limma_res_PA_clr <- run_topic_clr_limma(prep_PA$W_clean, prep_PA$meta_clean)
sig_limma_clr_PA  <- limma_res_PA_clr$res_df  %>% filter(adj_p_value_by_term < 0.05)
sig_limma_clr_PA 
```
```{r}
PA_limma_crl <- plot_limma(
  sig_limma_clr_PA ,
  topic_order = colnames(prep_PA$W_clean),
  title       = expression(bold("C") * " Limma") ,
  meta_names  = meta_names,
  param_order = param_order
)

PA_limma_crl
```


## RM + SHAP
```{r}
results_clr_PA <- run_rf_oob_clr(prep_PA$W_clean, prep_PA$meta_clean)
results_clr_PA
```
```{r}
#shap_dir_PA <- run_shap_direction(
#  rf_models  = results_clr_PA$rf_models,
#  meta_clean = prep_PA$meta_clean,
#  nsim       = 500
#)

#save so we don't need to re-run everytime
saveRDS(shap_dir_PA, file = "output/objects/PA/shap_dir_PA.rds")
shap_dir_PA <- readRDS("output/objects/PA/shap_dir_PA.rds")

shap_dir_PA
```

```{r}
# get the new names in the same order as param_order
new_order <- meta_names$new[match(param_order, meta_names$original)]

shap_dir_PA2<- shap_dir_PA %>%
  left_join(meta_names, by = c("Env" = "original")) %>%
  mutate(Env= coalesce(new, Env)) %>% 
  mutate(Env = factor(Env, levels = new_order))

PA_rfshap<-plot_importance_direction_dot(
  shap_dir    = shap_dir_PA2,
  rf_perf     = results_clr_PA$rf_perf,
  top_n       = 15,
  param_order = new_order,
  title       = expression(bold("D") * " RM + SHAP")
)

#ggsave("../output/plots/FL/FL_RFshap.png", p, width = 8, height = 6, units = "in",dpi=600)

PA_rfshap
```

#Summary plot
```{r}
PA_limma_crl_tight <- PA_limma_crl + theme(plot.margin = margin(0, -2, 0, 0))
PA_rfshap_tight    <- PA_rfshap    + theme(plot.margin = margin(0, 0, 0, -2))
bottom_right <- plot_grid(PA_limma_crl_tight, PA_rfshap_tight, ncol = 2,rel_widths = c(.8, 1.2))


right_col <- plot_grid(PA_meta_plt, bottom_right, nrow = 2,rel_heights = c(.8, 1.3))

#grab ht_PA as a grob if it's a ComplexHeatmap
ht_grob <- grid::grid.grabExpr(ComplexHeatmap::draw(ht_PA))
PA_summary <- plot_grid(ht_grob, right_col, 
                        ncol = 2,
                        align = "h",
                        axis = "tb",
                        rel_widths = c(1, 1.3))

ggsave("output/plots/PA/PA_summary.pdf", PA_summary, width = 18, height = 8, units = "in",dpi=600)
PA_summary
```

#____________
#### colors
```{r}
#colors
library(RColorBrewer)

# Get the 9-color Paired palette
paired_colors <- brewer.pal(8, "Paired")

# Create SSF colors
SSF_colors <- paired_colors
names(SSF_colors) <- paste0("SSF_", seq_along(SSF_colors))

# Create LSF colors 
my_palette <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", 
                "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", 
                "#fdae61")


LSF_colors <- my_palette 
names(LSF_colors) <- paste0("LSF_", seq_along(LSF_colors))

# Combine into one named vector
chl_scs <- c(SSF_colors, LSF_colors)

chl_scs

```

#**SSF**

```{r}
## same processing as the proks but we need to change the taxonomy labels 
process_phyloseq_EUK <- function(ps) {
  library(phyloseq)
  
  # Extract the three tables
  if (taxa_are_rows(ps)) {
    asv <- as.data.frame(t(otu_table(ps)))
  } else {
    asv <- as.data.frame(otu_table(ps))
  }
  
  tax  <- as.data.frame(tax_table(ps))
  meta <- as.data.frame(sample_data(ps))
  
  # Hand off to process_tables so we can clean if needed 
  return(process_tables_EUK(asv, tax, meta))
}

process_tables_EUK <- function(asv, tax, meta) {
  library(dplyr)
  
  tax$ASV <- rownames(tax)
  
  #we need asvs to be unique so we add a numer
  tax_dict <- tax %>%
    mutate(
      ASV_name = paste0(
        ifelse(is.na(Class), "[unknown Class]", Class),
        " (",
        ifelse(is.na(Order), "?", Order),
        ") ",
        row_number()
      )
    ) %>%
    ungroup()
  
  tax_dict$ASV <- factor(tax_dict$ASV, levels = colnames(asv))
  tax_dict <- tax_dict[order(tax_dict$ASV), ]
  
  colnames(asv) <- tax_dict$ASV_name
  asv <- as.data.frame(asv) %>% mutate(across(everything(), as.integer))
  
  return(list(
    counts   = as.data.frame(asv),
    tax_dict = tax_dict,
    meta     = meta
  ))
}
```


#2. Prep Data
```{r}
#this inlcudes counts, tax_dict, meta
SSF_results<-process_phyloseq_EUK(SSF_data)

#save the counts tables you will need for perplexity
write.csv(SSF_results$counts, "output/tables/runSSF/SSF_counts.csv")

#free up some memoery we don't need the FL_data anymore and we can call it in if we do 
rm(SSF_data)
gc()  

```

#3. Run LDA
### run LDA models 
```{r run-lda-models}
lda_varying_params_lists_SSF <-  list()
for (k in 1:20) {lda_varying_params_lists_SSF[[paste0("k",k)]] <- list(k = k)}

lda_models_SSF  <- 
  run_lda_models(
    data = SSF_results$counts,
    lda_varying_params_lists = lda_varying_params_lists_SSF,
    lda_fixed_params_list = list(method = "VEM"),
    dir = "../output/saved_models/runsSSF/",
    reset = FALSE,
    verbose = TRUE
  )

```

### align the topics based on different methods
```{r align-topics}

aligned_topics_product_SSF <- 
  align_topics(
    models = lda_models_SSF,
    method = "product") 

compute_number_of_paths(aligned_topics_product_SSF) %>% 
  plot_number_of_paths() + 
  ggtitle("Method: product")

aligned_topics_transport_SSF <- 
  align_topics(
    models = lda_models_SSF,
    method = "transport") 

compute_number_of_paths(aligned_topics_transport_SSF) %>% 
  plot_number_of_paths() + 
  ggtitle("Method: transport")

```

### top taxa alingment
```{r}
aligned_topics_transport_k <- 
  align_topics(
    models = lda_models_SSF[c(1,10)],
    method = "transport") 

p<-plot(aligned_topics_transport_k,
     add_leaves = TRUE, 
     label_topics = TRUE,
     n_features_in_leaves = 2,
     min_feature_prop = 0.01,
     leaves_text_size = 8,
     color_by="topic",
      model_name_repair_fun = function(x) str_remove_all(x, "k"))+
     theme_bw()+
    labs(x="")

#ggsave(p, filename = "PA_k10.png", height = 5, width = 9, units = "cm", scale = 1.5)

p
```

### perplexity (for diagnostics)
```{r}
SSF_perplexity = read.csv("perplexity/SSF_perplexity.csv",header=TRUE)

p<-ggplot(SSF_perplexity, aes(x = k, y = perplexity)) +
   geom_point() +
   geom_smooth(se = FALSE) +
   theme_bw()+
   labs(x = "Candidate number of topics", y = "Perplexity score")

p
```

##diagnostics figure
```{r}
figure <- make_figure2(aligned_topics_product = aligned_topics_product_SSF, 
                           aligned_topics_transport = aligned_topics_transport_SSF,
                       perplexity_df=SSF_perplexity,size_fraction="Small Chloroplast")


ggsave(figure, filename = "output/plots/SSF/diagnostics.png", height = 14, width = 21, units = "cm", scale = 1.5)

figure

```

Panel a: Paths track close to the diagonal at k=9, indicating a well-balanced model
Panel b: Coherence scores are mostly strong (0.5–0.8) at k=9, before degradation sets in at higher k
Panel c: Refinement score is near its peak at k=9, before variance becomes unstable
Panel d: k=9 sits just past the perplexity elbow, capturing most model improvement efficiently

### W and A
```{r}
W_SSF = as.data.frame(lda_models_SSF$k9$gamma)
colnames(W_SSF)<- sub("V", "", colnames(W_SSF))
W_SSF
```

```{r}
A_SSF <-as.data.frame(exp(lda_models_SSF$k9$beta))
A_df_SSF<-A_SSF ## to work on rare
A_SSF <- A_SSF %>% mutate(topic = row_number())

#change the ASV names 
A_long_SSF<- melt.data.frame(A_SSF, id.vars = "topic",variable_name="ASV")
A_long_SSF$Phyla<-sapply(strsplit(as.character(A_long_SSF$ASV)," "), "[", 1)
A_long_SSF$Order<-paste(sapply(strsplit(as.character(A_long_SSF$ASV)," "), "[", 2),
                    sapply(strsplit(as.character(A_long_SSF$ASV)," "), "[", 3),sep="_")
A_long_SSF$Order_raw<-paste(sapply(strsplit(as.character(A_long_SSF$Order),"_"), "[", 1))

#A_long_SSF<- A_long_SSF
#save(A_long_SSF,file="../output/tables/w_matrix/pa_tax")


A_long_SSF <- A_long_SSF %>% mutate(asv_type=case_when(value >= 0.1 ~ ">=10%", value < 0.1 ~ "<10%"))
A_long_SSF
```

#4.Run GoMDE
```{r}
res_SSF  <- run_gomde_markers(SSF_results$counts, SSF_results$tax_dict,W_SSF,A_long_SSF,  top_n = 25)

#save so you don't have to re-computer
saveRDS(res_SSF,"output/objects/SSF/res_SSF.rds")
res_SSF<-readRDS("output/objects/SSF/res_SSF.rds")

res_SSF
```

### overview
```{r}
#grab data from full set not top25 -- but still only SPECIALIST
mat_df <- res_SSF$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>% #sig enrichment 
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%      # specialists only
  ungroup() %>%
  left_join(SSF_results$tax_dict, by = c("ASV" = "ASV_name")) %>%
  filter(!is.na(Order)) %>%
  filter(!Order %in% c("unclass_plas")) %>%
  mutate(Class_clean = ifelse(Class %in% names(class_colors), Class, "Other")) %>%
  group_by(topic, Class, Class_clean, Order) %>%
  dplyr::summarise(n_asv = n(), .groups = "drop")


mat_df

## collapse orders we only see in 1 topic
min_asv <- 5  # orders with max < 3 across all topics become Other [Phylum]

# build matrix
# build full matrix first to know which orders are sparse
mat_full <- mat_df %>%
  group_by(Order, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order") %>%
  as.matrix()

# identify sparse orders
sparse_orders <- rownames(mat_full)[apply(mat_full, 1, max) < min_asv]

# collapse in mat_df
mat_df_collapsed <- mat_df %>%
  mutate(Order_plot = ifelse(Order %in% sparse_orders,
                           ifelse(Class_clean == "Other", 
                                  "Other (rare Class)",  # catch the Other Other case this is when we made phyla other
                                  paste0("Multiple ", Class_clean)),
                           Order)) %>% 
  group_by(topic, Class_clean, Order_plot) %>%
  dplyr::summarise(n_asv = sum(n_asv), .groups = "drop")

# rebuild matrix
mat <- mat_df_collapsed %>%
  group_by(Order_plot, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order_plot") %>%
  as.matrix()

mat <- mat[, order(as.numeric(colnames(mat)))]
colnames(mat) <- paste0("SSF_", colnames(mat))
mat <- mat[rowSums(mat) > 0, ]


order_class <- mat_df_collapsed %>%
  distinct(Order_plot, Class_clean) %>%
  filter(Order_plot %in% rownames(mat))

class_vec <- setNames(order_class$Class_clean, order_class$Order_plot)
class_vec <- class_vec[rownames(mat)]

class_colors_clean <- class_colors

# lighten phylum colors for Other rows
lighten <- function(hex, factor = 0.5) {
  rgb <- col2rgb(hex)
  rgb_light <- rgb + (255 - rgb) * factor
  rgb(rgb_light[1,], rgb_light[2,], rgb_light[3,], maxColorValue = 255)
}

# build annotation colors - named orders get full color, Other gets lightened
row_colors <- sapply(rownames(mat), function(ord) {
  class <- class_vec[ord]
  base_col <- class_colors_clean[class]
  if (grepl("^Multiple ", ord)) lighten(base_col, 0.5) else base_col
})

# for ComplexHeatmap we need the annotation to use the phylum color vector
# but mark Other rows differently with a border or pattern - simplest is
# just to show them as a named color per row
row_annot <- rowAnnotation(
  Class = class_vec,
  col = list(Class = class_colors_clean),
  show_annotation_name = FALSE,
  annotation_legend_param = list(Class = list(ncol = 1)),
  show_legend = TRUE  
)


# color scale
col_fun <- colorRamp2(c(0, max(mat)), c("white", "#2C7FB8"))

row_title <- rowAnnotation(
  label = anno_text("Order", 
                    rot = 90, 
                    just = "center",
                    gp = gpar(fontsize = 10, fontface = "bold")),
  width = unit(0.8, "cm")
)

row_name_cols <- ifelse(grepl("^Multiple", rownames(mat)), "grey50", "black")
row_name_fontface <- ifelse(grepl("^Multiple ", rownames(mat)), "italic", "plain")

ht_SSF <- Heatmap(mat,
  name = "ASV count",
  col = col_fun,
  left_annotation = row_annot,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_dend = TRUE,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 8,
                      col = row_name_cols,
                      fontface = row_name_fontface),
  column_names_gp = gpar(fontsize = 10),
  column_title = expression(bold("A") * " Fingerprint Taxa"),
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  rect_gp = gpar(col = "grey90", lwd = 0.5),
  width = unit(10, "cm"),
  row_names_max_width = max_text_width(rownames(mat), gp = gpar(fontsize = 8)),
  layer_fun = function(j, i, x, y, width, height, fill) {
    v <- pindex(mat, i, j)
    grid.text(ifelse(v > 0, as.character(v), ""), x, y,
              gp = gpar(fontsize = 7, col = "grey20"))
  },
  row_title = "Order",
  row_title_gp = gpar(fontsize = 10, fontface = "bold"),
  row_title_side = "right"
) 

ht_SSF

```


### top markers summary
```{r}
# swarm data FIRST (defines topic order)
df_swarm <- res_SSF$top_markers %>%
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%        # pecialists only
  ungroup() %>%
   mutate(
    topic  = factor(topic),
    Class = ifelse(Class %in% names(class_colors), Class, "Other")
  )

# pie composition per topic
pie_df <- df_swarm %>%
  count(topic, Class, name = "n") %>%
  group_by(topic) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  select(topic, Class, prop) %>%
  pivot_wider(names_from = Class, values_from = prop, values_fill = 0)

# CONTROL PIE LOCATION (global annotation band)
topic_levels <- levels(df_swarm$topic)
pie_df <- pie_df %>%
  mutate(x = as.numeric(factor(topic, levels = topic_levels)))

y_max <- max(df_swarm$log2FC, na.rm = TRUE)
pie_df$y <- y_max + 1.5   # <-- you can set this to whatever you want

pie_cols <- intersect(names(pie_df), names(class_colors))

## topic levels
topic_levels <- df_swarm %>%
  distinct(topic) %>%
  arrange(as.numeric(sub(".*_", "", topic))) %>%
  pull(topic)

df_swarm$topic <- factor(df_swarm$topic, levels = topic_levels)
pie_df$topic  <- factor(pie_df$topic,  levels = topic_levels)
pie_df$x <- as.numeric(pie_df$topic)


p_swarm_pie_SSF <- ggplot(df_swarm, aes(x = topic, y = log2FC, color = Class)) + #size = log2FC
  geom_quasirandom(width = 0.3, alpha = .8,stroke = 0.9 ) +
  scatterpie::geom_scatterpie(
    data = pie_df,
    aes(x = x, y = y),
    cols = pie_cols,
    pie_scale = 3,
    color = "white",
    linewidth = 0.3) +
  scale_color_manual(values = class_colors) +
  scale_fill_manual(values = class_colors) +
  guides(fill = "none", color = guide_legend(ncol = 4,byrow = TRUE)) + #size = guide_legend(ncol = 1)
  #guides(fill = "none")+
  theme_bw() +
  labs(x = "Topic", y = "GoM-DE log2FC - Top25", color = "Class")+ #size = "log2FC"
  coord_fixed(ratio = 0.6, ylim = c(0, y_max + 2))+
  theme(
    panel.grid.major.x = element_blank(),
    legend.title = element_text(angle = 90),
    legend.spacing.x = unit(0.1, "mm"),
    legend.spacing.y = unit(0.1, "mm"),
     legend.key.size = unit(1, "mm"),
    legend.position = "bottom",
    legend.direction = "horizontal")

p_swarm_pie_SSF
```

```{r}
asv_topic_sig_SSF <- res_SSF$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>%            # enriched only
  left_join(distinct(SSF_results$tax_dict, ASV_name, Class),
            by = c("ASV" = "ASV_name")) %>%
  mutate(
    Class = ifelse(Class %in% names(class_colors), Class, "Other"),
    Class = factor(Class, levels = names(class_colors))
  ) %>%
  group_by(ASV) %>%
  summarise(
    Class = dplyr::first(Class),
    n_topics = n_distinct(topic),
    max_log2FC = max(log2FC),
    .groups = "drop"
  )

df_hm_SSF <- asv_topic_sig_SSF %>%
  count(Class, n_topics)
print(sum(df_hm_SSF$n))

pheat_SSF<-ggplot(df_hm_SSF, aes(x = factor(n_topics), y = Class, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  scale_fill_gradient(
    low = "#f2e5ff",
    high = "#5b2a86"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Number of topics where ASV is significantly enriched",
    y = "Class",
    fill = "ASV count",
    title = paste0(
      "Distribution of marker ASVs (n=",
      nrow(asv_topic_sig_SSF),
      ")"
     )
  )
pheat_SSF
```

```{r}
figure <-  ggpubr::ggarrange(pheat_SSF,p_swarm_pie_SSF, ncol = 2, nrow = 1,widths = c(.4,.6))

figure<-annotate_figure(figure,
                top = text_grob("Small Euks", color = "black", face = "bold", size = 14))

ggsave(file="output/plots/SSF/gomde_SSF_pie.pdf", figure, width = 12, height = 8, units = "in",dpi=600)

figure
```

#5.Run Env Analysis

### topic proportion across samples
```{r}
## cluster by similarity
w_den<-as.dendrogram(as.hclust(agnes(W_SSF, diss=F, method =  "ward")),hang=.5)

#data based on the clustering above 
W_SSF2<-W_SSF
W_SSF2$Sample_num<-row.names(W_SSF2)
W_long_SSF<- melt.data.frame( W_SSF2, id.vars="Sample_num", variable_name="topic")

plota_SSF<-W_long_SSF %>%
      ggplot(aes(x=factor(Sample_num,level=labels(w_den)), value, fill=topic))+
      geom_col(color=NA,width=1,linewidth=0)+
      scale_fill_brewer(palette="Paired")+
      labs(y="Subcommunity Proportion",x=paste0("Samples (n=", dim(W_SSF2)[1],")"),fill="SC#",
           title=expression(bold("C") * " SSF Chloroplast"))+
      theme_bw()+
    scale_x_discrete(labels = NULL, breaks = NULL) + 
    scale_y_continuous(breaks = seq(0, 1, 0.1),expand = c(0, 0))+
      theme(axis.title.y = element_text(size = 10),
             plot.title = element_text(face = "bold", size = 12),
          axis.text.y = element_text(size = 10),
          legend.key.height= unit(.05, 'in'),
        legend.key.width= unit(.05, 'in'),
        legend.box.spacing = unit(0.01, "pt"))

ggsave("output/plots/SSF/SSF_structure.pdf", plota_SSF, width = 7, height = 2.5, units = "in",dpi=600)

plota_SSF

```

```{r}
SSF_meta<-data.frame(SSF_results$meta, check.names = FALSE)

#PLOTS B-D: legend bars
#get sample infor from main clean table
sample_info<-SSF_meta[,c('Sample_num','new_depths_mdh2','CTD_TEMP','SEASON','YEAR','lake')]
#drop duplicates - because each OTU gets a value so we have many reps 
colnames(sample_info)<-c('sample_id','depth','CTD_TEMP','SEASON','YEAR','lake')
#reorder to match the order of the bar plots 
sample_info<-sample_info %>% arrange(sample_id,levels=labels(w_den))
sample_info$sample_id<- factor(sample_info$sample_id,level=labels(w_den))

###make the bar plots 
lake<-ggplot(data=sample_info, aes(x=sample_id, fill=lake)) + 
  scale_fill_manual(values=lake_colors)+
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

season<-ggplot(data=sample_info, aes(x=sample_id, fill=SEASON)) + 
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=season_colors)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

TEMP<-ggplot(data=sample_info, aes(x=sample_id, fill=CTD_TEMP)) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_gradient(low = "blue", high = "red")+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")
  
YEAR<-ggplot(data=sample_info, aes(x=sample_id, fill=factor(YEAR))) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=year_colors)+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

depth<-ggplot(data=sample_info, aes(x=sample_id, fill=depth)) + 
  scale_fill_manual(values=depth_colors_w)+
  geom_bar(width = 1,size=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

wx1=0.072
wx2=0.73

### value definitions
names_depth=c("spr:","I","E","T","H","smr:","E","D","T","H")

colors_depth= c(season_colors["Spring"],depth_colors_w["Iso_Top"], depth_colors_w["InvS_EPI"],
          depth_colors_w["InvS_TRM"], depth_colors_w["InvS_HYP"], season_colors["Summer"],
          depth_colors_w["Strat_EPI"],depth_colors_w["Strat_DCL"],depth_colors_w["Strat_TRM"],
          depth_colors_w["Strat_HYP"])

#change dynamically
label_start   <- wx1 + wx2 + 0.01
label_spacing <- 0.03

season_locx <- label_start + c(0, 2)                            * label_spacing  # 2 labels
lake_locx   <- label_start + c(0, 1, 2, 3, 4)                  * label_spacing  # 5 labels
strata_locx <- label_start + c(0, 0.03, 0.04, 0.05, 0.06, 0.09, 0.12, 0.13, 0.14, 0.15)
year_locx <- label_start + c(0, 0.02, 0.035, 0.05, 0.065, 0.08, 0.095, 0.11, 0.125) #8 labels
temp_locx   <- label_start + c(0, 1, 2)                        * label_spacing  # 3 labels


#gather all the plots
SSF_structure<- ggdraw()+ theme_void()+
 draw_plot(plota_SSF,x = 0, y = 0.25, width = 0.878, height = 0.75)+
 draw_plot(season,x = wx1, y= 0.2, width = wx2, height = 0.05)+
  draw_plot(lake,x = wx1, y=0.15, width = wx2, height = 0.05)+
  draw_plot(depth,x = wx1, y = 0.1, width =wx2, height = 0.05)+
  draw_plot(YEAR,x = wx1, y = 0.05, width =wx2, height = 0.05)+
  draw_plot(TEMP,x = wx1, y = 0, width =wx2, height = 0.05)+
  draw_plot_label(x = c(0.01,0.01,0.01,0.01,0.01), y = c(0.25,0.2,0.15,0.1,0.05), 
                     label =  c("Season","Lake","Depth","Year","Temp"),
                     hjust = 0, size = 7)+
  draw_plot_label(x=c(season_locx,lake_locx, strata_locx, year_locx, temp_locx), #temp
                  y=c(rep(0.25,2), #season x 2
                      rep(0.2,5), #lake x5
                      rep(0.15,10), #depth x 10
                      rep(0.1,9), #year x 9
                      rep(0.05,3)), #temp x3
                  label=c('Spring','Summer',
                          "SU","MI","HU","ER","ON",
                          names_depth,
                          "201","2","3","4","5","6","7","8","9",
                          'Low','High', 'NA'), 
                  color=c(season_colors["Spring"], season_colors["Summer"],
                          "#92809E", "#8cabea", "#64baaa","#FFC86A","#ee7061",
                          colors_depth,
                          "gray",'red','#A6CEE3','#B2DF8A',"brown","violet","purple","blue","black",
                          "blue","red","gray"),
                  hjust = 0, size = 6 )


  
  
ggsave("output/plots/SSF/SSF_sample_membership.pdf", SSF_structure, width = 7, height = 3, units = "in",dpi=600)
  
SSF_structure
```




### topic location
```{r}
#meta long
SSF_meta_long<- melt.data.frame(subset(SSF_meta,strat_status != "missing_data"), id.vars="Sample_num", variable_name="variable")
W_long_meta_SSF<-full_join(SSF_meta_long, W_long_SSF, by = "Sample_num")

p_season<-subset(W_long_meta_SSF,(variable == "SEASON") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(season_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=season_colors)+
      theme_bw()+
      labs(fill="Season",title="",y="Relative contribution",x= "Subcommunity")

p_season

p_lake<-subset(W_long_meta_SSF,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      theme_bw()+
      labs(fill="Lake",title="",y="",x= "")
p_lake

p_layer<-subset(W_long_meta_SSF,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      theme_classic()+
      labs(fill="Layer",title="",y="",x= "")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )



p_layer

p_year<-subset(W_long_meta_SSF,(variable == "YEAR") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(year_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=year_colors)+
      theme_bw()+
      labs(fill="Year",title="",y="",x= "")

p_year

p_status<-subset(W_long_meta_SSF,(variable == "strat_status2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill= value.x))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=strat_colors)+
      labs(x="",y="", fill="Strat. Type")+
      theme_bw()

p_status

#sample prop - how much does this sc represent overall 
hold<-subset(W_long_meta_SSF, variable == "YEAR")  #can be any variable 
prop<-hold %>% group_by(topic) %>% dplyr::summarise(prop=round((sum(value.y)/sum(hold$value.y))*100))

p_dist<-ggplot(prop, aes(x=topic, y=prop, fill=topic)) +
  geom_col() +
  geom_text(aes(label=paste0(round(prop, 1), "%")), 
            vjust=-0.5, 
            size=3, 
            fontface="bold") +  
  scale_fill_brewer(palette="Paired") +
  labs(x="", y="% of all sample distribution") +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))+
  theme(legend.position="none", 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(),
        theme(axis.title=element_text(size=8, face="bold")))


figure = ggpubr::ggarrange(p_season, p_status, p_layer, p_lake, p_year,p_dist, ncol = 3, nrow = 2)

figure<-ggpubr::annotate_figure(figure, 
                        top = ggpubr::text_grob("SSF EUK: Topic Distribution across samples", color = "black", 
                                        face = "bold", size = 12))

ggsave("output/plots/SSF/SSF_membership.pdf", figure, width = 9, height = 5, units = "in",dpi=600)

figure

```


```{r}

p_lake<-subset(W_long_meta_SSF,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Lake",title=expression(bold("B") * " Meta Overview"), 
           y="Relative contribution",x= "Subcommunity")
p_lake

p_layer<-subset(W_long_meta_SSF,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Layer",title="",y="Relative contribution",x= "Subcommunity")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )

SSF_meta_plt<- ggpubr::ggarrange(p_lake, p_layer, ncol = 2, nrow = 1)
SSF_meta_plt
```

##prep meta
```{r}
prep_SSF   <- prep_topic_meta(W_SSF, SSF_meta)
```
## Limma
```{r}
limma_res_SSF_clr <- run_topic_clr_limma(prep_SSF$W_clean, prep_SSF$meta_clean)
sig_limma_clr_SSF  <- limma_res_SSF_clr$res_df  %>% filter(adj_p_value_by_term < 0.05)
sig_limma_clr_SSF 
```

```{r}
SSF_limma_crl <- plot_limma(
  sig_limma_clr_SSF ,
  topic_order = colnames(prep_SSF$W_clean),
  title       = expression(bold("C") * " Limma") ,
  meta_names  = meta_names,
  param_order = param_order
)

SSF_limma_crl
```


## RM + SHAP
```{r}
results_clr_SSF <- run_rf_oob_clr(prep_SSF$W_clean, prep_SSF$meta_clean)
results_clr_SSF
```
```{r}
shap_dir_SSF <- run_shap_direction(
  rf_models  = results_clr_SSF$rf_models,
  meta_clean = prep_SSF$meta_clean,
  nsim       = 500
)

#save so we don't need to re-run everytime
saveRDS(shap_dir_SSF, file = "output/objects/SSF/shap_dir_SSF.rds")
shap_dir_SSF <- readRDS("output/objects/SSF/shap_dir_SSF.rds")

shap_dir_SSF
```

```{r}
# get the new names in the same order as param_order
new_order <- meta_names$new[match(param_order, meta_names$original)]

shap_dir_SSF2<- shap_dir_SSF %>%
  left_join(meta_names, by = c("Env" = "original")) %>%
  mutate(Env= coalesce(new, Env)) %>% 
  mutate(Env = factor(Env, levels = new_order))

SSF_rfshap<-plot_importance_direction_dot(
  shap_dir    = shap_dir_SSF2,
  rf_perf     = subset(results_clr_SSF$rf_perf, Topic !="6"), ##6 is dropped from Limma so drop here too it has low R2
  top_n       = 15,
  param_order = new_order,
  title       = expression(bold("D") * " RM + SHAP")
)

#ggsave("../output/plots/FL/FL_RFshap.png", p, width = 8, height = 6, units = "in",dpi=600)

SSF_rfshap

```

#Summary plot
```{r}
SSF_limma_crl_tight <- SSF_limma_crl + theme(plot.margin = margin(0, -2, 0, 0))
SSF_rfshap_tight    <- SSF_rfshap    + theme(plot.margin = margin(0, 0, 0, -2))
bottom_right <- plot_grid(SSF_limma_crl_tight, SSF_rfshap_tight, ncol = 2,rel_widths = c(.8, 1.2))


right_col <- plot_grid(SSF_meta_plt, bottom_right, nrow = 2,rel_heights = c(.8, 1.3))

#grab ht_SSF as a grob if it's a ComplexHeatmap
ht_grob <- grid::grid.grabExpr(ComplexHeatmap::draw(ht_SSF))
SSF_summary <- plot_grid(ht_grob, right_col, 
                        ncol = 2,
                        align = "h",
                        axis = "tb",
                        rel_widths = c(1, 1.3))

ggsave("output/plots/SSF/SSF_summary.pdf", SSF_summary, width = 18, height = 8, units = "in",dpi=600)
SSF_summary
```



#____________
#**LSF**

#2. Prep Data
```{r}
#this inlcudes counts, tax_dict, meta
LSF_results<-process_phyloseq_EUK(LSF_data)

#save the counts tables you will need for perplexity
write.csv(LSF_results$counts, "output/tables/runLSF/LSF_counts.csv")

#free up some memoery we don't need the FL_data anymore and we can call it in if we do 
rm(LSF_data)
gc()  

```

#3. Run LDA
### run LDA models 
```{r run-lda-models}
lda_varying_params_lists_LSF <-  list()
for (k in 1:20) {lda_varying_params_lists_LSF[[paste0("k",k)]] <- list(k = k)}

lda_models_LSF  <- 
  run_lda_models(
    data = LSF_results$counts,
    lda_varying_params_lists = lda_varying_params_lists_LSF,
    lda_fixed_params_list = list(method = "VEM"),
    dir = "../output/saved_models/runsLSF/",
    #reset = TRUE,
    #verbose = TRUE
  )

```


### align the topics based on different methods
```{r align-topics}

aligned_topics_product_LSF <- 
  align_topics(
    models = lda_models_LSF,
    method = "product") 

compute_number_of_paths(aligned_topics_product_LSF) %>% 
  plot_number_of_paths() + 
  ggtitle("Method: product")

aligned_topics_transport_LSF <- 
  align_topics(
    models = lda_models_LSF,
    method = "transport") 

compute_number_of_paths(aligned_topics_transport_LSF) %>% 
  plot_number_of_paths() + 
  ggtitle("Method: transport")

```

### top taxa alingment
```{r}
aligned_topics_transport_k <- 
  align_topics(
    models = lda_models_LSF[c(1,13)],
    method = "transport") 

p<-plot(aligned_topics_transport_k,
     add_leaves = TRUE, 
     label_topics = TRUE,
     n_features_in_leaves = 2,
     min_feature_prop = 0.01,
     leaves_text_size = 8,
     color_by="topic",
      model_name_repair_fun = function(x) str_remove_all(x, "k"))+
     theme_bw()+
    labs(x="")

#ggsave(p, filename = "PA_k10.png", height = 5, width = 9, units = "cm", scale = 1.5)

p
```

### perplexity (for diagnostics)
```{r}
LSF_perplexity = read.csv("perplexity/LSF_perplexity.csv",header=TRUE)

p<-ggplot(LSF_perplexity, aes(x = k, y = perplexity)) +
   geom_point() +
   geom_smooth(se = FALSE) +
   theme_bw()+
   labs(x = "Candidate number of topics", y = "Perplexity score")

p
```

##diagnostics figure
```{r}
figure <- make_figure2(aligned_topics_product = aligned_topics_product_LSF, 
                           aligned_topics_transport = aligned_topics_transport_LSF,
                       perplexity_df=LSF_perplexity,size_fraction="Large Chloroplast")


ggsave(figure, filename = "output/plots/LSF/diagnostics.png", height = 14, width = 21, units = "cm", scale = 1.5)

figure

```

Panel a: Paths track close to the diagonal at k=11
Panel b: Coherence scores are mostly strong (0.5–0.8) at k=11, before degradation sets in at higher k
Panel c: Refinement score has plateaued by k=11, with little gain from adding more topics
Panel d: k=11 sits at the flattening point of the perplexity curve, capturing most model improvement efficiently
And for interpretation less is better 


### W and A
```{r}
W_LSF = as.data.frame(lda_models_LSF$k11$gamma)
colnames(W_LSF)<- sub("V", "", colnames(W_LSF))
W_LSF
```

```{r}
A_LSF <-as.data.frame(exp(lda_models_LSF$k11$beta))
A_df_LSF<-A_LSF ## to work on rare
A_LSF <- A_LSF %>% mutate(topic = row_number())

#change the ASV names 
A_long_LSF<- melt.data.frame(A_LSF, id.vars = "topic",variable_name="ASV")
A_long_LSF$Phyla<-sapply(strsplit(as.character(A_long_LSF$ASV)," "), "[", 1)
A_long_LSF$Order<-paste(sapply(strsplit(as.character(A_long_LSF$ASV)," "), "[", 2),
                    sapply(strsplit(as.character(A_long_LSF$ASV)," "), "[", 3),sep="_")
A_long_LSF$Order_raw<-paste(sapply(strsplit(as.character(A_long_LSF$Order),"_"), "[", 1))

#A_long_LSF<- A_long_LSF
#save(A_long_LSF,file="../output/tables/w_matrix/pa_tax")


A_long_LSF <- A_long_LSF %>% mutate(asv_type=case_when(value >= 0.1 ~ ">=10%", value < 0.1 ~ "<10%"))
A_long_LSF
```

#4.Run GoMDE
```{r}
#res_LSF  <- run_gomde_markers(LSF_results$counts, LSF_results$tax_dict,W_LSF,A_long_LSF,  top_n = 25)

#save so you don't have to re-computer
#saveRDS(res_LSF,"output/objects/LSF/res_LSF.rds")
res_LSF<-readRDS("output/objects/LSF/res_LSF.rds")

res_LSF
```

### overview
```{r}
#grab data from full set not top25 -- but still only SPECIALIST
mat_df <- res_LSF$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>% #sig enrichment 
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%      # specialists only
  ungroup() %>%
  left_join(LSF_results$tax_dict, by = c("ASV" = "ASV_name")) %>%
  filter(!is.na(Order)) %>%
  filter(!Order %in% c("unclass_plas")) %>%
  mutate(Class_clean = ifelse(Class %in% names(class_colors), Class, "Other")) %>%
  group_by(topic, Class, Class_clean, Order) %>%
  dplyr::summarise(n_asv = n(), .groups = "drop")


mat_df

## collapse orders we only see in 1 topic
min_asv <- 5  # orders with max < 3 across all topics become Other [Phylum]

# build matrix
# build full matrix first to know which orders are sparse
mat_full <- mat_df %>%
  group_by(Order, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order") %>%
  as.matrix()

# identify sparse orders
sparse_orders <- rownames(mat_full)[apply(mat_full, 1, max) < min_asv]

# collapse in mat_df
mat_df_collapsed <- mat_df %>%
  mutate(Order_plot = ifelse(Order %in% sparse_orders,
                           ifelse(Class_clean == "Other", 
                                  "Other (rare Class)",  # catch the Other Other case this is when we made phyla other
                                  paste0("Multiple ", Class_clean)),
                           Order)) %>% 
  group_by(topic, Class_clean, Order_plot) %>%
  dplyr::summarise(n_asv = sum(n_asv), .groups = "drop")

# rebuild matrix
mat <- mat_df_collapsed %>%
  group_by(Order_plot, topic) %>%
  summarise(n_asv = sum(n_asv), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = topic, values_from = n_asv, values_fill = 0) %>%
  tibble::column_to_rownames("Order_plot") %>%
  as.matrix()

mat <- mat[, order(as.numeric(colnames(mat)))]
colnames(mat) <- paste0("LSF_", colnames(mat))
mat <- mat[rowSums(mat) > 0, ]


order_class <- mat_df_collapsed %>%
  distinct(Order_plot, Class_clean) %>%
  filter(Order_plot %in% rownames(mat))

class_vec <- setNames(order_class$Class_clean, order_class$Order_plot)
class_vec <- class_vec[rownames(mat)]

class_colors_clean <- class_colors

# lighten phylum colors for Other rows
lighten <- function(hex, factor = 0.5) {
  rgb <- col2rgb(hex)
  rgb_light <- rgb + (255 - rgb) * factor
  rgb(rgb_light[1,], rgb_light[2,], rgb_light[3,], maxColorValue = 255)
}

# build annotation colors - named orders get full color, Other gets lightened
row_colors <- sapply(rownames(mat), function(ord) {
  class <- class_vec[ord]
  base_col <- class_colors_clean[class]
  if (grepl("^Multiple ", ord)) lighten(base_col, 0.5) else base_col
})

# for ComplexHeatmap we need the annotation to use the phylum color vector
# but mark Other rows differently with a border or pattern - simplest is
# just to show them as a named color per row
row_annot <- rowAnnotation(
  Class = class_vec,
  col = list(Class = class_colors_clean),
  show_annotation_name = FALSE,
  annotation_legend_param = list(Class = list(ncol = 1)),
  show_legend = TRUE  
)


# color scale
col_fun <- colorRamp2(c(0, max(mat)), c("white", "#2C7FB8"))

row_title <- rowAnnotation(
  label = anno_text("Order", 
                    rot = 90, 
                    just = "center",
                    gp = gpar(fontsize = 10, fontface = "bold")),
  width = unit(0.8, "cm")
)

row_name_cols <- ifelse(grepl("^Multiple", rownames(mat)), "grey50", "black")
row_name_fontface <- ifelse(grepl("^Multiple ", rownames(mat)), "italic", "plain")

ht_LSF <- Heatmap(mat,
  name = "ASV count",
  col = col_fun,
  left_annotation = row_annot,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_dend = TRUE,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 8,
                      col = row_name_cols,
                      fontface = row_name_fontface),
  column_names_gp = gpar(fontsize = 10),
  column_title = expression(bold("A") * " Fingerprint Taxa"),
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  rect_gp = gpar(col = "grey90", lwd = 0.5),
  width = unit(10, "cm"),
  row_names_max_width = max_text_width(rownames(mat), gp = gpar(fontsize = 8)),
  layer_fun = function(j, i, x, y, width, height, fill) {
    v <- pindex(mat, i, j)
    grid.text(ifelse(v > 0, as.character(v), ""), x, y,
              gp = gpar(fontsize = 7, col = "grey20"))
  },
  row_title = "Order",
  row_title_gp = gpar(fontsize = 10, fontface = "bold"),
  row_title_side = "right"
) 

ht_LSF

```

## top markers summary
```{r}
# swarm data FIRST (defines topic order)
df_swarm <- res_LSF$top_markers %>%
  group_by(ASV) %>%
  filter(n_distinct(topic) == 1) %>%        # specialists only
  ungroup() %>%
   mutate(
    topic  = factor(topic),
    Class = ifelse(Class %in% names(class_colors), Class, "Other")
  )

# pie composition per topic
pie_df <- df_swarm %>%
  count(topic, Class, name = "n") %>%
  group_by(topic) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  select(topic, Class, prop) %>%
  pivot_wider(names_from = Class, values_from = prop, values_fill = 0)

# CONTROL PIE LOCATION (global annotation band)
topic_levels <- levels(df_swarm$topic)
pie_df <- pie_df %>%
  mutate(x = as.numeric(factor(topic, levels = topic_levels)))

y_max <- max(df_swarm$log2FC, na.rm = TRUE)
pie_df$y <- y_max + 1.5   # <-- you can set this to whatever you want

pie_cols <- intersect(names(pie_df), names(class_colors))

## topic levels
topic_levels <- df_swarm %>%
  distinct(topic) %>%
  arrange(as.numeric(sub(".*_", "", topic))) %>%
  pull(topic)

df_swarm$topic <- factor(df_swarm$topic, levels = topic_levels)
pie_df$topic  <- factor(pie_df$topic,  levels = topic_levels)
pie_df$x <- as.numeric(pie_df$topic)


p_swarm_pie_LSF <- ggplot(df_swarm, aes(x = topic, y = log2FC, color = Class)) + #size = log2FC
  geom_quasirandom(width = 0.3, alpha = .8,stroke = 0.9 ) +
  scatterpie::geom_scatterpie(
    data = pie_df,
    aes(x = x, y = y),
    cols = pie_cols,
    pie_scale = 3,
    color = "white",
    linewidth = 0.3) +
  scale_color_manual(values = class_colors) +
  scale_fill_manual(values = class_colors) +
  guides(fill = "none", color = guide_legend(ncol = 4,byrow = TRUE)) + #size = guide_legend(ncol = 1)
  #guides(fill = "none")+
  theme_bw() +
  labs(x = "Topic", y = "GoM-DE log2FC - Top25", color = "Class")+ #size = "log2FC"
  coord_fixed(ratio = 0.6, ylim = c(0, y_max + 2))+
  theme(
    panel.grid.major.x = element_blank(),
    legend.title = element_text(angle = 90),
    legend.spacing.x = unit(0.1, "mm"),
    legend.spacing.y = unit(0.1, "mm"),
     legend.key.size = unit(1, "mm"),
    legend.position = "bottom",
    legend.direction = "horizontal")

p_swarm_pie_LSF
```

```{r}
asv_topic_sig_LSF <- res_LSF$gomde %>%
  filter(lfsr < 0.05, log2FC > 0) %>%            # enriched only
  left_join(distinct(LSF_results$tax_dict, ASV_name, Class),
            by = c("ASV" = "ASV_name")) %>%
  mutate(
    Class = ifelse(Class %in% names(class_colors), Class, "Other"),
    Class = factor(Class, levels = names(class_colors))
  ) %>%
  group_by(ASV) %>%
  summarise(
    Class = dplyr::first(Class),
    n_topics = n_distinct(topic),
    max_log2FC = max(log2FC),
    .groups = "drop"
  )

df_hm_LSF <- asv_topic_sig_LSF %>%
  count(Class, n_topics)
print(sum(df_hm_LSF$n))

pheat_LSF<-ggplot(df_hm_LSF, aes(x = factor(n_topics), y = Class, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  scale_fill_gradient(
    low = "#f2e5ff",
    high = "#5b2a86"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Number of topics where ASV is significantly enriched",
    y = "Class",
    fill = "ASV count",
    title = paste0(
      "Distribution of marker ASVs (n=",
      nrow(asv_topic_sig_LSF),
      ")"
     )
  )
pheat_LSF
```

```{r}
figure <-  ggpubr::ggarrange(pheat_LSF,p_swarm_pie_LSF, ncol = 2, nrow = 1,widths = c(.4,.6))

figure<-annotate_figure(figure,
                top = text_grob("Large Euks", color = "black", face = "bold", size = 14))

ggsave(file="output/plots/LSF/gomde_LSF_pie.pdf", figure, width = 12, height = 8, units = "in",dpi=600)

figure
```

#5.Run Env Analysis

### topic proportion across samples
```{r}
## cluster by similarity
w_den<-as.dendrogram(as.hclust(agnes(W_LSF, diss=F, method =  "ward")),hang=.5)

#data based on the clustering above 
W_LSF2<-W_LSF
W_LSF2$Sample_num<-row.names(W_LSF2)
W_long_LSF<- melt.data.frame( W_LSF2, id.vars="Sample_num", variable_name="topic")

plota_LSF<-W_long_LSF %>%
      ggplot(aes(x=factor(Sample_num,level=labels(w_den)), value, fill=topic))+
      geom_col(color=NA,width=1,linewidth=0)+
      scale_fill_brewer(palette="Paired")+
      labs(y="Subcommunity Proportion",x=paste0("Samples (n=", dim(W_LSF2)[1],")"),fill="SC#",
           title=expression(bold("D") * " LSF Chloroplast"))+
      theme_bw()+
    scale_x_discrete(labels = NULL, breaks = NULL) + 
    scale_y_continuous(breaks = seq(0, 1, 0.1),expand = c(0, 0))+
      theme(axis.title.y = element_text(size = 10),
             plot.title = element_text(face = "bold", size = 12),
          axis.text.y = element_text(size = 10),
          legend.key.height= unit(.05, 'in'),
        legend.key.width= unit(.05, 'in'),
        legend.box.spacing = unit(0.01, "pt"))

ggsave("output/plots/LSF/LSF_structure.pdf", plota_LSF, width = 7, height = 2.5, units = "in",dpi=600)

plota_LSF

```

```{r}
LSF_meta<-data.frame(LSF_results$meta, check.names = FALSE)

#PLOTS B-D: legend bars
#get sample infor from main clean table
sample_info<-LSF_meta[,c('Sample_num','new_depths_mdh2','CTD_TEMP','SEASON','YEAR','lake')]
#drop duplicates - because each OTU gets a value so we have many reps 
colnames(sample_info)<-c('sample_id','depth','CTD_TEMP','SEASON','YEAR','lake')
#reorder to match the order of the bar plots 
sample_info<-sample_info %>% arrange(sample_id,levels=labels(w_den))
sample_info$sample_id<- factor(sample_info$sample_id,level=labels(w_den))

###make the bar plots 
lake<-ggplot(data=sample_info, aes(x=sample_id, fill=lake)) + 
  scale_fill_manual(values=lake_colors)+
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

season<-ggplot(data=sample_info, aes(x=sample_id, fill=SEASON)) + 
  geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=season_colors)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

TEMP<-ggplot(data=sample_info, aes(x=sample_id, fill=CTD_TEMP)) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_gradient(low = "blue", high = "red")+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")
  
YEAR<-ggplot(data=sample_info, aes(x=sample_id, fill=factor(YEAR))) + 
geom_bar(width = 1,linewidth=0)+ theme_void()+
  scale_fill_manual(values=year_colors)+
theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
  strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

depth<-ggplot(data=sample_info, aes(x=sample_id, fill=depth)) + 
  scale_fill_manual(values=depth_colors_w)+
  geom_bar(width = 1,size=0)+ theme_void()+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
    strip.text.x = element_blank(),legend.position="none")+xlab("") + ylab("")

wx1=0.072
wx2=0.73

### value definitions
names_depth=c("spr:","I","E","T","H","smr:","E","D","T","H")

colors_depth= c(season_colors["Spring"],depth_colors_w["Iso_Top"], depth_colors_w["InvS_EPI"],
          depth_colors_w["InvS_TRM"], depth_colors_w["InvS_HYP"], season_colors["Summer"],
          depth_colors_w["Strat_EPI"],depth_colors_w["Strat_DCL"],depth_colors_w["Strat_TRM"],
          depth_colors_w["Strat_HYP"])

#change dynamically
label_start   <- wx1 + wx2 + 0.01
label_spacing <- 0.03

season_locx <- label_start + c(0, 2)                            * label_spacing  # 2 labels
lake_locx   <- label_start + c(0, 1, 2, 3, 4)                  * label_spacing  # 5 labels
strata_locx <- label_start + c(0, 0.03, 0.04, 0.05, 0.06, 0.09, 0.12, 0.13, 0.14, 0.15)
year_locx <- label_start + c(0, 0.02, 0.035, 0.05, 0.065, 0.08, 0.095, 0.11, 0.125) #8 labels
temp_locx   <- label_start + c(0, 1, 2)                        * label_spacing  # 3 labels


#gather all the plots
LSF_structure<- ggdraw()+ theme_void()+
 draw_plot(plota_LSF,x = 0, y = 0.25, width = 0.878, height = 0.75)+
 draw_plot(season,x = wx1, y= 0.2, width = wx2, height = 0.05)+
  draw_plot(lake,x = wx1, y=0.15, width = wx2, height = 0.05)+
  draw_plot(depth,x = wx1, y = 0.1, width =wx2, height = 0.05)+
  draw_plot(YEAR,x = wx1, y = 0.05, width =wx2, height = 0.05)+
  draw_plot(TEMP,x = wx1, y = 0, width =wx2, height = 0.05)+
  draw_plot_label(x = c(0.01,0.01,0.01,0.01,0.01), y = c(0.25,0.2,0.15,0.1,0.05), 
                     label =  c("Season","Lake","Depth","Year","Temp"),
                     hjust = 0, size = 7)+
  draw_plot_label(x=c(season_locx,lake_locx, strata_locx, year_locx, temp_locx), #temp
                  y=c(rep(0.25,2), #season x 2
                      rep(0.2,5), #lake x5
                      rep(0.15,10), #depth x 10
                      rep(0.1,9), #year x 9
                      rep(0.05,3)), #temp x3
                  label=c('Spring','Summer',
                          "SU","MI","HU","ER","ON",
                          names_depth,
                          "201","2","3","4","5","6","7","8","9",
                          'Low','High', 'NA'), 
                  color=c(season_colors["Spring"], season_colors["Summer"],
                          "#92809E", "#8cabea", "#64baaa","#FFC86A","#ee7061",
                          colors_depth,
                          "gray",'red','#A6CEE3','#B2DF8A',"brown","violet","purple","blue","black",
                          "blue","red","gray"),
                  hjust = 0, size = 6 )


  
  
ggsave("output/plots/LSF/LSF_sample_membership.pdf", LSF_structure, width = 7, height = 3, units = "in",dpi=600)
  
LSF_structure
```




### topic location
```{r}
#meta long
LSF_meta_long<- melt.data.frame(subset(LSF_meta,strat_status != "missing_data"), id.vars="Sample_num", variable_name="variable")
W_long_meta_LSF<-full_join(LSF_meta_long, W_long_LSF, by = "Sample_num")

p_season<-subset(W_long_meta_LSF,(variable == "SEASON") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(season_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=season_colors)+
      theme_bw()+
      labs(fill="Season",title="",y="Relative contribution",x= "Subcommunity")

p_season

p_lake<-subset(W_long_meta_LSF,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      theme_bw()+
      labs(fill="Lake",title="",y="",x= "")
p_lake

p_layer<-subset(W_long_meta_LSF,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      theme_classic()+
      labs(fill="Layer",title="",y="",x= "")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )



p_layer

p_year<-subset(W_long_meta_LSF,(variable == "YEAR") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(year_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=year_colors)+
      theme_bw()+
      labs(fill="Year",title="",y="",x= "")

p_year

p_status<-subset(W_long_meta_LSF,(variable == "strat_status2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill= value.x))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=strat_colors)+
      labs(x="",y="", fill="Strat. Type")+
      theme_bw()

p_status

#sample prop - how much does this sc represent overall 
hold<-subset(W_long_meta_LSF, variable == "YEAR")  #can be any variable 
prop<-hold %>% group_by(topic) %>% dplyr::summarise(prop=round((sum(value.y)/sum(hold$value.y))*100))

p_dist<-ggplot(prop, aes(x=topic, y=prop, fill=topic)) +
  geom_col() +
  geom_text(aes(label=paste0(round(prop, 1), "%")), 
            vjust=-0.5, 
            size=3, 
            fontface="bold") +  
  scale_fill_brewer(palette="Paired") +
  labs(x="", y="% of all sample distribution") +
  theme_classic() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))+
  theme(legend.position="none", 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(),
        theme(axis.title=element_text(size=8, face="bold")))


figure = ggpubr::ggarrange(p_season, p_status, p_layer, p_lake, p_year,p_dist, ncol = 3, nrow = 2)

figure<-ggpubr::annotate_figure(figure, 
                        top = ggpubr::text_grob("LSF EUK: Topic Distribution across samples", color = "black", 
                                        face = "bold", size = 12))

ggsave("output/plots/LSF/LSF_membership.pdf", figure, width = 9, height = 5, units = "in",dpi=600)

figure

```


```{r}

p_lake<-subset(W_long_meta_LSF,(variable == "lake") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(lake_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=lake_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Lake",title=expression(bold("B") * " Meta Overview"), 
           y="Relative contribution",x= "Subcommunity")
p_lake

p_layer<-subset(W_long_meta_LSF,(variable == "new_depths_mdh2") & (value.y>0)) %>%
      ggplot(aes(topic, value.y, fill=factor(value.x,levels=names(depth2_colors))))+
      geom_col(position="fill")+
      #geom_col()+
      scale_fill_manual(values=depth2_colors)+
      scale_alpha_manual(values = c("yes" = 1, "no" = 0.3), guide = "none") +
      theme_classic()+
      labs(fill="Layer",title="",y="Relative contribution",x= "Subcommunity")+
      theme(
        legend.position="right",  # Adjust if needed
        legend.key.size=unit(0.5, "cm"),  # Reduce legend key size
        legend.text=element_text(size=8),  # Smaller legend text
        legend.spacing.x=unit(0.2, "cm")  # Reduce spacing between legend items
      )

LSF_meta_plt<- ggpubr::ggarrange(p_lake, p_layer, ncol = 2, nrow = 1)
LSF_meta_plt
```



##prep meta
```{r}
prep_LSF   <- prep_topic_meta(W_LSF, LSF_meta)
```
## Limma
```{r}
limma_res_LSF_clr <- run_topic_clr_limma(prep_LSF$W_clean, prep_LSF$meta_clean)
sig_limma_clr_LSF  <- limma_res_LSF_clr$res_df  %>% filter(adj_p_value_by_term < 0.05)
sig_limma_clr_LSF 
```

```{r}
LSF_limma_crl <- plot_limma(
  sig_limma_clr_LSF ,
  topic_order = colnames(prep_LSF$W_clean),
  title       = expression(bold("C") * " Limma") ,
  meta_names  = meta_names,
  param_order = param_order
)

LSF_limma_crl
```


## RM + SHAP
```{r}
results_clr_LSF <- run_rf_oob_clr(prep_LSF$W_clean, prep_LSF$meta_clean)
results_clr_LSF
```
```{r}
# shap_dir_LSF <- run_shap_direction(
#   rf_models  = results_clr_LSF$rf_models,
#   meta_clean = prep_LSF$meta_clean,
#   nsim       = 500
# )

#save so we don't need to re-run everytime
saveRDS(shap_dir_LSF, file = "output/objects/LSF/shap_dir_LSF.rds")
shap_dir_LSF <- readRDS("output/objects/LSF/shap_dir_LSF.rds")

shap_dir_LSF
```

```{r}
# get the new names in the same order as param_order
new_order <- meta_names$new[match(param_order, meta_names$original)]

shap_dir_LSF2<- shap_dir_LSF %>%
  left_join(meta_names, by = c("Env" = "original")) %>%
  mutate(Env= coalesce(new, Env)) %>% 
  mutate(Env = factor(Env, levels = new_order))

LSF_rfshap<-plot_importance_direction_dot(
  shap_dir    = shap_dir_LSF2,
  rf_perf     = results_clr_LSF$rf_perf,
  top_n       = 15,
  param_order = new_order,
  title       = expression(bold("D") * " RM + SHAP")
)

#ggsave("../output/plots/FL/FL_RFshap.png", p, width = 8, height = 6, units = "in",dpi=600)

LSF_rfshap

```
#Summary Plot
```{r}
LSF_limma_crl_tight <- LSF_limma_crl + theme(plot.margin = margin(0, -2, 0, 0))
LSF_rfshap_tight    <- LSF_rfshap    + theme(plot.margin = margin(0, 0, 0, -2))
bottom_right <- plot_grid(LSF_limma_crl_tight, LSF_rfshap_tight, ncol = 2,rel_widths = c(.8, 1.2))


right_col <- plot_grid(LSF_meta_plt, bottom_right, nrow = 2,rel_heights = c(.8, 1.3))

#grab ht_LSF as a grob if it's a ComplexHeatmap
ht_grob <- grid::grid.grabExpr(ComplexHeatmap::draw(ht_LSF))
LSF_summary <- plot_grid(ht_grob, right_col, 
                        ncol = 2,
                        align = "h",
                        axis = "tb",
                        rel_widths = c(1, 1.3))

ggsave("output/plots/LSF/LSF_summary.pdf", LSF_summary, width = 18, height = 8, units = "in",dpi=600)
LSF_summary
```


#____________
# All blocks
## Samples used
```{r}
library(ggtext)
```

```{r}
station_order=c("SU17M","SU08M","SU01M","SUFE","SU18","SU15","SU12","SU06",
                "HU54M","HU45M","HU15M","HUFE",
                "MI41M","MI27M","MI18M","MI40","MI50B","MIFO",
                "ER91M","ER78M","ER15M",
                "ON33M","ON41","ON55M","ON68B","ON64B")

plot_sample_heatmap <- function(X_meta, letter, name) {
  
  summary <- X_meta %>% 
    group_by(STATION_ID, SEASON, YEAR) %>% 
    dplyr::summarise(total_samples = n_distinct(Sample_num), .groups = "drop")
  
  summary$season_year <- paste(summary$SEASON, summary$YEAR, sep = "_")
  summary <- summary[, c("STATION_ID", "season_year", "total_samples")]
  
  station_order <- sort(unique(summary$STATION_ID))
  
  p <- ggplot(summary, 
              aes(x = season_year, y = factor(STATION_ID, levels = rev(station_order)), fill = total_samples)) +
    geom_tile() +
    coord_fixed() +
    theme_classic() +
    theme(axis.text.x = element_markdown(angle = 90)) +
    scale_fill_gradient(low = "lightgreen", high = "darkgreen") +
    geom_text(aes(label = total_samples), color = "white", size = 3) +
    labs(fill = "# Samples", y = "Station ID", x = "Season",
         title = bquote(bold(.(letter)) ~ .(paste0(name, " n=", sum(summary$total_samples)))))
  
  return(p)
}

p_fl<- plot_sample_heatmap(FL_meta, "A", "Free Living Prokaryotes")
p_pa<- plot_sample_heatmap(PA_meta, "B", "Particle Associated Prokaryotes")
p_ssf<- plot_sample_heatmap(SSF_meta, "C", "Small Chloroplast")
p_lsf<- plot_sample_heatmap(LSF_meta, "D", "Large Chloroplast")


all_samples<-ggpubr::ggarrange(p_fl, p_pa, p_ssf, p_lsf, 
                                 ncol = 2, nrow = 2)

ggsave("output/plots/ALL/ALL_samples_used.pdf", all_samples, width = 12, height = 12, units = "in",dpi=600)
all_samples

```


## Structure summary 
```{r}
all_strcuture<-ggpubr::ggarrange(FL_structure,PA_structure,
                                 SSF_structure,LSF_structure,
                                 ncol = 2, nrow = 2)

ggsave("output/plots/ALL/ALL_structure.pdf", all_strcuture, width = 14, height = 6, units = "in",dpi=600)
all_strcuture
```

## GoMDE
```{r}
# Grab FL GoM-DE heatmap as grob
ht_grob_FL <- grid::grid.grabExpr({
  ComplexHeatmap::draw(ht_FL2)
  
  grid.text("Phylum", 
            x = unit(0.05, "npc"),  # tweak x
            y = unit(0.03, "npc"),  # tweak y — top
            gp = gpar(fontsize = 9, fontface = "bold"),
            just = c("left", "bottom"))
  
  grid.text("Order", 
            x = unit(0.75, "npc"),  # tweak x — right side
            y = unit(0.03, "npc"),   # vertical center
            gp = gpar(fontsize = 9, fontface = "bold"),
            just = c("right", "bottom"))
})


pheat2<-pheat+labs(title=expression(bold("B")*""))
p_swarm_pie2<-p_swarm_pie+labs(title=expression(bold("C")*""))

# Stack B and C on the right
right_col <- plot_grid(
  pheat2,  # panel B
  p_swarm_pie2,        # panel C
  nrow = 2,
  rel_heights = c(1, 1)  # B is shorter, C taller
)

# Combine A (left) and B/C stack (right)
gomde_fl <- plot_grid(
  ht_grob_FL, right_col,
  ncol = 2,
  align = "h",
  axis = "tb",
  rel_widths = c(1, 1)  # GoM-DE wider, adjust to taste
)

ggsave("output/plots/FL/GomDE_summary.pdf", gomde_fl, width = 14, height = 12, units = "in",dpi=600)

gomde_fl
```



## Env associations
```{r}
# Combine all blocks
limma_all <- bind_rows(
  sig_limma_clr_FL  %>% mutate(Block = "FL"),
  sig_limma_clr_PA  %>% mutate(Block = "PA"),
  sig_limma_clr_SSF %>% mutate(Block = "SSF"),
  sig_limma_clr_LSF %>% mutate(Block = "LSF")
) %>%
  mutate(Block = factor(Block, levels = c("FL", "PA", "SSF", "LSF")))

# Summarise per block × env variable
limma_summary <- limma_all %>%
  group_by(Block, term) %>%
  summarise(
    n_sig = n(),  # number of SCs with significant association
    mean_t = mean(t, na.rm = TRUE),  # dominant direction
    .groups = "drop"
  ) %>%
  mutate(
    direction_label = ifelse(mean_t > 0, "Positive", "Negative")
  ) %>%
  left_join(meta_names, by = c("term" = "original")) %>%
  mutate(term = coalesce(new, term))

# Plot
p_limma_summary <- ggplot(limma_summary,
                          aes(x = Block, y = term)) +
  geom_point(aes(size = n_sig, color = direction_label), alpha = 0.9) +
  scale_color_manual(values = c("Positive" = "#E65100",
                                 "Negative" = "#1565C0"),
                     name = "Dominant\ndirection") +
  scale_size_continuous(range = c(3, 10),
                        name = "Number of \n sig. SCs") +
  scale_x_discrete(limits = c("FL", "PA", "SSF", "LSF")) +
  labs(x = NULL, y = NULL,
       title = expression(bold("A") * " Limma Observed Associations")) +
  theme_bw(base_size = 11) +
  theme(panel.grid.major = element_line(color = "grey92"),
        panel.grid.minor = element_blank())

p_limma_summary 
```



```{r}
# Combine all blocks
rf_perf_all <- bind_rows(
  results_clr_FL$rf_perf  %>% mutate(Block = "FL"),
  results_clr_PA$rf_perf  %>% mutate(Block = "PA"),
  results_clr_SSF$rf_perf %>% mutate(Block = "SSF"),
  results_clr_LSF$rf_perf %>% mutate(Block = "LSF")
) %>%
  mutate(Block = factor(Block, levels = c("FL", "PA", "SSF", "LSF")))

# Block means
rf_means <- rf_perf_all %>%
  group_by(Block) %>%
  summarise(mean_R2 = mean(R2))

rf_means 

# Plot
p_q2 <- ggplot(rf_perf_all, aes(y = R2, x = Block, color = Block)) +
  geom_jitter(height = 0.15, size = 2.5, alpha = 0.7) +
  geom_crossbar(data = rf_means,
                aes(y = mean_R2, ymin = mean_R2, ymax = mean_R2),
                width = 0.5, linewidth = 0.5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_discrete(limits = c("FL", "PA", "SSF", "LSF")) +
  scale_color_manual(values = sf_colors2) +
  labs(x = NULL, title = expression(bold("B") * " RF Prediction Power"),
       y = expression(Q^2 ~ "(OOB " * R^2 * ")")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank())

p_q2

```
```{r}
# Combine all blocks
shap_all <- bind_rows(
  shap_dir_FL  %>% mutate(Block = "FL"),
  shap_dir_PA  %>% mutate(Block = "PA"),
  shap_dir_SSF %>% mutate(Block = "SSF"),
  shap_dir_LSF %>% mutate(Block = "LSF")
) %>%
  mutate(Block = factor(Block, levels = c("FL", "PA", "SSF", "LSF")))

shap_summary <- shap_all %>%
  mutate(direction_num = case_when(
    direction == "+" ~ 1,
    direction == "-" ~ -1,
    TRUE ~ NA_real_
  )) %>%
  group_by(Block, Env) %>%
  summarise(
    mean_imp  = mean(imp_scaled, na.rm = TRUE),
    prop_pos  = mean(direction_num == 1, na.rm = TRUE),
    n_topics  = n(),
    .groups = "drop"
  ) %>%
  mutate(
    dom_dir = case_when(
      prop_pos >= 0.70 ~ "+",
      prop_pos <= 0.30 ~ "−",
      TRUE ~ ""  # no dominant direction, leave blank
    )
  )

shap_plot <- shap_summary %>% 
  filter(show) %>%
  mutate(direction_label = ifelse(dir_color == "#E65100", 
                                   "Positive", "Negative")) %>%
  left_join(meta_names, by = c("Env" = "original")) %>%
  mutate(Env = coalesce(new, Env)) %>%
  mutate(Env = droplevels(factor(Env, levels = new_order)))

env_order <- shap_plot %>%
  group_by(Env) %>%
  summarise(total_imp = sum(mean_imp)) %>%
  arrange(total_imp) %>%
  pull(Env)

p_drivers <- ggplot(shap_plot, 
                    aes(x = Block, y = Env)) +
  geom_point(aes(color = direction_label, size = mean_imp), alpha = 0.9) +
  #geom_text(aes(label = dom_dir), 
  #          color = "white", size = 3.5, fontface = "bold") +
  scale_color_manual(values = c("Positive" = "#E65100", 
                                 "Negative" = "#1565C0"),
                     name = "Dominant\ndirection") +
  scale_size_continuous(range = c(5, 14), 
                        name = "Mean importance\n(scaled)") +
  scale_x_discrete(limits = c("FL", "PA", "SSF", "LSF")) +
  scale_y_discrete(limits = env_order) +
  labs(x = NULL, y = NULL, title = expression(bold("C") * " Predictive Importance")) +
  theme_bw(base_size = 11) +
  theme(panel.grid.major = element_line(color = "grey92"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5))

p_drivers 
```


```{r}
p<-ggpubr::ggarrange(p_limma_summary,p_q2, p_drivers,
                                 ncol = 3, nrow = 1, widths = c(1,.75,1))

ggsave("output/plots/ALL/meta_stats.pdf", p, width = 15, height = 6, units = "in",dpi=600)
p
```


## Distribution and entropy
### ASV
```{r}
ASV_distribution<- function(name, df, cutoff){
  #filter
  df_clean<- df %>% 
    filter (value > cutoff) %>% 
    group_by(ASV) %>% 
    dplyr::summarise(n_topics=n_distinct(topic)) %>% 
    ungroup() %>% 
    arrange(n_topics) %>% 
    mutate(sf=name)
  
  p<-ggplot(df_clean, aes(x=n_topics)) + 
    geom_bar()+
    theme_bw()+
    labs(x="found in this # of SCs", y="ASVs", title=paste(name, cutoff))+
    scale_x_continuous(breaks = seq(min(df_clean$n_topics), max(df_clean$n_topics), by = 1))
  
  print(p)
  
   return(df_clean)
}

#ASV_distribution("PA",A_long_PA, 0.0001)
ASV_distribution("SSF_Chloro",A_long_SSF, 0.0001)
```

```{r}
fl_asvs_dist<-ASV_distribution("FL_Prok",A_long_FL, 0.0001)
pa_asvs_dist<-ASV_distribution("PA_Prok",A_long_PA, 0.0001)

ssf_asvs_dist<-ASV_distribution("SSF_Chloro",A_long_SSF, 0.0001)
lsf_asvs_dist<-ASV_distribution("LSF_Chloro",A_long_LSF, 0.0001)

join<-rbind(fl_asvs_dist,pa_asvs_dist,ssf_asvs_dist,lsf_asvs_dist)

#bacuse not all are observed
max_values <- c(FL_Prok = 10, PA_Prok = 10, SSF_Chloro = 9, LSF_Chloro = 11)

ecdf_data <- join %>%
  group_by(sf, n_topics) %>%
  dplyr::summarise(n = n(), .groups = "drop") %>%
  group_split(sf) %>%
  map_dfr(function(df_sf) {
    sf_name <- unique(df_sf$sf)
    df_sf %>%
      complete(n_topics = 1:max_values[[sf_name]], fill = list(n = 0)) %>%
      arrange(n_topics) %>%
      mutate(
        sf = sf_name,
        cdf = cumsum(n) / sum(n)
      )}) %>% 
  mutate(sf=factor(sf,levels=names(max_values)))

points <- ecdf_data %>%
  filter(n_topics %in% c(1,9))
points_adjusted <- points %>%
  filter(n_topics %in% c(9)) %>%
  mutate(cdf_adj = cdf + c(-.02,-.04,-.06,-.08)[as.numeric(factor(sf))])

p_asvs<- ggplot(ecdf_data, aes(x = n_topics, y = cdf, color = sf)) +
  geom_line(size = .5) +
  geom_point(size = 3, alpha=.7) +
  theme_bw() +
  labs(
    x = "# of SCs containing each ASV",
    y = "Cumulative Fraction of ASVs",
    color = "Dataset\n(p>0.0001)",
    title = expression(bold("B") ~ "")) + #ASV Distribution Across SCs
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = c(0.8, 0.3))+
  scale_x_continuous(breaks = seq(min(join$n_topics), 13, by = 1))+
  scale_y_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_color_manual(values=sf_colors)+
  geom_text(
    data = subset(points,n_topics ==1),
    aes(label = n),
    hjust = 1.2, vjust = 0.5,
    size = 3,
    show.legend = FALSE)+
  geom_text(
      data = points_adjusted,
      aes(x = n_topics, y = cdf_adj, label = n, color = sf),
      hjust = -0.1,
      size = 3,
      show.legend = FALSE)+
  annotate("text", 
         x = .7, y = 0.6, 
         label = "n=", 
         size = 3, fontface = "bold", color = "black", hjust = 0)+
  guides(color="none")

p_asvs
```

### Samples
```{r}
sample_distribution<- function(name, df, cutoff){
  
  # reshape
  df$sample <- rownames(df) 
  
  df_long <- df %>%
    pivot_longer(
      cols = where(is.numeric),    
      names_to = "topic",
      values_to = "value"
    )
  
  # Reorder columns
  df_long <- df_long %>% select(sample, topic, value)
  
  # filter
  df_clean <- df_long %>% 
    filter(value > cutoff) %>% 
    group_by(sample) %>% 
    dplyr::summarise(n_topics = n_distinct(topic)) %>% 
    ungroup() %>% 
    arrange(n_topics) %>% 
    mutate(sf = name)
  
  p <- ggplot(df_clean, aes(x = n_topics)) + 
    geom_bar() +
    theme_bw() +
    labs(x = "found in this # of SCs", y = "samples", title = paste(name, cutoff)) +
    scale_x_continuous(breaks = seq(min(df_clean$n_topics), max(df_clean$n_topics), by = 1))
  
  print(p)
  
  return(df_clean)
  
}

sample_distribution("FL",W_FL, 0.01)
#sample_distribution("LSF",W_LSF, 0.01)
```

```{r}
fl_sample_dist<-sample_distribution("FL_Prok",W_FL, 0.01)
pa_sample_dist<-sample_distribution("PA_Prok",W_PA, 0.01)

ssf_sample_dist<-sample_distribution("SSF_Chloro",W_SSF, 0.01)
lsf_sample_dist<-sample_distribution("LSF_Chloro",W_LSF, 0.01)

join<-rbind(fl_sample_dist,pa_sample_dist,ssf_sample_dist,lsf_sample_dist)

#bacuse not all are observed
max_values <- c(FL_Prok = 10, PA_Prok = 10, SSF_Chloro = 9, LSF_Chloro = 11)

ecdf_data <- join %>%
  group_by(sf, n_topics) %>%
  dplyr::summarise(n = n(), .groups = "drop") %>%
  group_split(sf) %>%
  map_dfr(function(df_sf) {
    sf_name <- unique(df_sf$sf)
    df_sf %>%
      complete(n_topics = 1:max_values[[sf_name]], fill = list(n = 0)) %>%
      arrange(n_topics) %>%
      mutate(
        sf = sf_name,
        cdf = cumsum(n) / sum(n)
      )}) %>% 
  mutate(sf=factor(sf,levels=names(max_values)))

p_samples<- ggplot(ecdf_data, aes(x = n_topics, y = cdf, color = sf)) +
  geom_line(size = .5) +
  geom_point(size = 3, alpha=.7) +
  theme_bw() +
  labs(
    x = "# of SCs observed in each sample",
    y = "Cumulative Fraction of Samples",
    color = "Dataset\n", #(p>0.01)
   title = expression(bold("C") ~ " ")) + #SC distribution across samples
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = c(0.8, 0.3))+
  scale_x_continuous(breaks = seq(min(join$n_topics), 13, by = 1))+
  scale_y_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_color_manual(values=sf_colors)


p_samples
```


### Entropy
Higher entropy in a region (FL or PA) → May indicate an environment with greater microbial interaction, more functional redundancy, or dynamic shifts in community structure.

Lower entropy → Suggests strong environmental filtering, where certain conditions select for specific microbial groups, leading to more stable and distinct sub-communities.

```{r}
library(entropy)  # Install with install.packages("entropy") if needed
```

```{r}
# Function to compute Shannon entropy for each row (document)
compute_entropy <- function(topic_matrix) {
  apply(topic_matrix, 1, function(probabilities) {
    probabilities <- probabilities[probabilities > 0]  # Remove zero probabilities (to avoid log(0))
    -sum(probabilities * log(probabilities, base = 2))  # Shannon entropy formula
  })
}

# Compute entropy for each dataset
FL_entropy <- compute_entropy(W_FL)
PA_entropy <- compute_entropy(W_PA)
SSF_entropy <- compute_entropy(W_SSF)
LSF_entropy <- compute_entropy(W_LSF)

# Create a combined data frame for plotting
entropy_df <- data.frame(
  Entropy = c(FL_entropy, PA_entropy, SSF_entropy, LSF_entropy),
  Dataset = rep(c("FL_Prok", "PA_Prok", "SSF_Chloro", "LSF_Chloro"), 
                c(length(FL_entropy), length(PA_entropy), length(SSF_entropy), length(LSF_entropy)))
)

entropy_df$Dataset<-factor(entropy_df$Dataset, levels=c("FL_Prok", "PA_Prok", "SSF_Chloro", "LSF_Chloro"))

# Compute summary statistics (mean and standard deviation)
summary_df <- entropy_df %>%
  group_by(Dataset) %>%
  dplyr::summarise(
    MeanEntropy = mean(Entropy),
    SdEntropy = sd(Entropy)
  )

# Convert summary stats to text labels
summary_df <- summary_df %>%
  mutate(Label = paste0("Mean: ", round(MeanEntropy, 2), "\nSD: ", round(SdEntropy, 2)))


# Plot using ggplot with facet_wrap() for subplots
p_ent<-ggplot(entropy_df, aes(x=Entropy, fill=Dataset, color=Dataset)) +
  geom_density(alpha=0.6) +
  labs(x="Shannon Entropy Across Dataset", y="Density",
       title = expression(bold("D") ~ "")) + #Entropy of Sample Subcommunity Composition
  theme_minimal() +
  theme(
    legend.position = c(0.9, 0.7),
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold"))+
  scale_fill_manual(values=sf_colors) + 
  scale_color_manual(values=sf_colors) +
  #facet_wrap(~ Dataset)+
  guides(color="none", fill="none")+
  geom_text(data=subset(summary_df,Dataset == "FL_Prok"), 
            aes(x=2.1, y=.8, label=Label), color="#1f78b4", size=4, inherit.aes=FALSE)+
  geom_text(data=subset(summary_df,Dataset == "PA_Prok"), 
            aes(x=1, y=.75, label=Label), color="#b15928", size=4, inherit.aes=FALSE)+
  geom_text(data=subset(summary_df,Dataset == "SSF_Chloro"), 
            aes(x=0.2, y=.65, label=Label), color="#33a02c", size=4, inherit.aes=FALSE)+
  geom_text(data=subset(summary_df,Dataset == "LSF_Chloro"), 
            aes(x=2.6, y=.3, label=Label), color="#00441b", size=4, inherit.aes=FALSE)
p_ent
```
```{r}
# Perform KS tests
ks_FL_PA <- ks.test(FL_entropy, PA_entropy)  # FL vs. PA
ks_SSF_LSF <- ks.test(SSF_entropy, LSF_entropy)  # SSF vs. LSF

# Print results
cat("Kolmogorov-Smirnov Test Results:\n")
cat("FL vs. PA: D =", ks_FL_PA$statistic, ", p-value =", ks_FL_PA$p.value, "\n")
cat("SSF vs. LSF: D =", ks_SSF_LSF$statistic, ", p-value =", ks_SSF_LSF$p.value, "\n")

```
### gather
```{r}
figure <-  ggpubr::ggarrange(p_asvs,p_samples,p_ent, ncol = 3, nrow = 1,
                             align = "h",heights = c(1, 1, 1),hjust = -0.1)

figure<-annotate_figure(figure,
                top = text_grob("", color = "black", face = "bold", size = 14))

ggsave(file=paste("output/plots/ALL/distribution_and_entropy.pdf",sep=''), figure,
       width = 15, height = 5, units = "in",dpi=1000)

figure
```

## GCA
```{r}
library(geigen)
library(pracma)
library(VGAM)
library(compositions)
```


### combine
We can only run with samples found in all datastes. 
A decision I am making is to make FL the master set, it was our first set and PA samples where processed based on FL.
Then SSF and LSF come from both FL and PA so wherever we dont have obs is not that we didn't sample its that there are no chloroplast reads so I am creating an absent topic.

- master sample list is the FL samples
- combine blocks, fill missing topics with 1 PER BLOCK

```{r}
combine_all <- function(){
  "wrap in a function so we don't mess up our main tables"
  
  # helper to prefix numeric topic columns -- 
  add_prefix <- function(df, prefix){
    df %>% rename_with(~ paste0(prefix, "_", .), where(is.numeric))
  }
  
  # FL
  FL <- W_FL %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var = "Sample_num") %>%
    add_prefix("FL")
  
  # SSF
  SSF <- W_SSF %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var = "Sample_num") %>%
    add_prefix("SSF")
  
  # PA ID lookup
  PA_ids <- PA_meta %>% select(Sample_num, ssf_id)
  
  # PA
  PA <- W_PA %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var = "Sample_num") %>%
    add_prefix("PA") %>%
    left_join(PA_ids, by = "Sample_num") %>%
    select(-Sample_num) %>%
    dplyr::rename(Sample_num = ssf_id)
  
  # LSF
  LSF <- W_LSF %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var = "Sample_num") %>%
    add_prefix("LSF") %>%
    left_join(PA_ids, by = "Sample_num") %>%
    select(-Sample_num) %>%
    dplyr::rename(Sample_num = ssf_id)
  
  # combine all
  all_wide <- FL %>%
    left_join(PA,  by = "Sample_num") %>%
    left_join(SSF, by = "Sample_num") %>%
    left_join(LSF, by = "Sample_num")
  
  return(all_wide)
}

all_wide <- combine_all()
all_wide

```


```{r}
add_missing_topic <- function(df, prefix) {
  # all columns that start with "PA_", "SSF_", "LSF_", etc.
  topic_cols <- grep(paste0("^", prefix, "_"), names(df), value = TRUE)
  
  # don't treat existing *_0 as part of the set (in case you call it twice)
  topic_cols <- setdiff(topic_cols, paste0(prefix, "_0"))
  
  new_col <- paste0(prefix, "_0")
  
  df %>%
    rowwise() %>%
    mutate(
      !!new_col := if_else(
        all(is.na(c_across(all_of(topic_cols)))),  # all topics NA for this block?
        1,                                         # yes → in "missing topic"
        0                                          # no  → not missing
      )
    ) %>%
    ungroup() %>%
    mutate(
      across(all_of(topic_cols), ~ replace_na(.x, 0))
    )
}
```

```{r}
all_wide2 <- all_wide %>%
  add_missing_topic("PA") %>%
  add_missing_topic("SSF") %>%
  add_missing_topic("LSF")

#clean
all_wide2 <- all_wide2 %>% 
  filter(PA_0 !=1) %>% #this is the PA fraction not in FL - this will also drop samples from LSF
  select(-PA_0)
  

all_wide2
```

```{r}
print(sum(all_wide2$SSF_0)) #samples from FL without euks
print(sum(all_wide2$LSF_0)) #samples from FL+PA without euks
```

### prep blocks
```{r}
# Standardize environmental data (M)
M_FL_scaled <- FL_meta %>% 
  filter(Sample_num %in% all_wide2$Sample_num) %>% 
  select(all_of(keep)) %>%         ## keep is defined in FL prep_meta
  mutate(across(everything(), ~replace_na(.x, 0))) %>% 
  mutate(across(everything(), ~ .x + 1)) %>% 
  mutate(across(everything(), scale)) %>% 
  mutate(across(everything(), as.numeric)) 
M_FL_scaled <- as.data.frame(M_FL_scaled)
M_FL_scaled
```
```{r}
FL_block  <- all_wide2 %>% select(starts_with("FL_"))  %>% as.matrix()
PA_block  <- all_wide2 %>% select(starts_with("PA_"))  %>% as.matrix()
SSF_block <- all_wide2 %>% select(starts_with("SSF_")) %>% as.matrix()
LSF_block <- all_wide2 %>% select(starts_with("LSF_")) %>% as.matrix()

add_rownames <- function(transformed, original){
  rownames(transformed) <- rownames(original)
  transformed
}

dataset_list <- list(
  FL  = add_rownames(ilr(FL_block),  FL_block),
  PA  = add_rownames(ilr(PA_block),  PA_block),
  SSF = add_rownames(ilr(SSF_block), SSF_block),
  LSF = add_rownames(ilr(LSF_block), LSF_block),
  Env = M_FL_scaled
)
```


### GCA function 
What are the major ecological axes that simultaneously structure prokaryotic (FL/PA) and eukaryotic (SSF/LSF) topics AND correlate with environmental gradients?
```{r}
#the goal is to prioritize variation shared across ALL blocks not just high noise within one block 

compute_gca <- function(dataset_list, nb_components = 2, scale = FALSE, lambda=10) {
  # Compute generalized canonical analysis
  # dataset_list: list of datasets to analyze
  # nb_components: number of components to compute
  
  ### Step 1: append all datasets
  dim_list <- unlist(lapply(dataset_list, ncol))
  
  if (scale) {
    ### Usually a bit better to scale the data
    dataset_list <- lapply(dataset_list, scale)
  }else{
    dataset_list <- lapply(dataset_list, as.matrix)
  }
  full_data <- bind_cols(dataset_list)
  #Sigma is the global covariance matrix across all variables in all blocks.
  Sigma  <- cov(full_data)
  #Sigma0 = covariance only within each block, NO cross-block covariance <- represents where blocks are independent 
  Sigma0 <- matrix(0, nrow = sum(dim_list), ncol = sum(dim_list))
  for (i in 1:length(dataset_list)){
    #### fill in the covariance matrix (block wise)
    a = ifelse(i==1, 1, sum(dim_list[1:(i-1)])+1)
    b = sum(dim_list[1:i])
    Sigma0[a:b, a:b] <- cov(dataset_list[[i]])
  }
  
  ### Step 2: compute the generalized eigenvalue problem
  sq_0 = pracma::sqrtm(Sigma0 + lambda * diag(rep(1, ncol(Sigma0))), kmax=100)  
  #goal is to make blocsk comparable to each other by minimizing their own variance 

  Sigma0_sqrt_inv <- sq_0$Binv
  
  #solve the eigne this is where cross-block covariance is high AND within-block covariance is low.
  res <- eigen(Sigma0_sqrt_inv  %*% Sigma  %*% Sigma0_sqrt_inv)
 
   ####  compute the canonical vectors
  A = Sigma0_sqrt_inv%*% res$vectors[,1:nb_components]
  results <- list()
  results[["canonical_vectors"]] <- list()

  for (i in seq_along(dataset_list)) {
    a <- ifelse(i == 1, 1, sum(dim_list[1:(i-1)]) + 1)
    b <- sum(dim_list[1:i])
    results[["canonical_vectors"]][[i]] <- A[a:b,]
  }
  
  # Initialize another list inside results
  results[["canonical_variates"]] <- list()
  
  for (i in seq_along(dataset_list)) {
    a <- ifelse(i == 1, 1, sum(dim_list[1:(i-1)]) + 1)
    b <- sum(dim_list[1:i])
    results[["canonical_variates"]][[i]] <- dataset_list[[i]] %*% A[a:b,]
  }
  
  results[['correlations']] = res$values
  results[['prop_variance_explained']]= res$values/sum(res$values)
  return(results)
}

```

```{r}
gca_out <- compute_gca(
  dataset_list,
  nb_components = 5,
  scale = TRUE,   # usually reasonable since these are probabilities
)
names(gca_out)
```

#### Loadings
```{r}
prop_df <- tibble(
  component = seq_along(gca_out$prop_variance_explained),
  prop = gca_out$prop_variance_explained
)

ggplot(prop_df, aes(x = factor(component), y = prop)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            vjust = -0.5, size = 3) +
  labs(x = "Component", y = "Proportion of Shared Variance Explained",
       title = "GCA Scree Plot") +
  theme_minimal()
```
```{r}
groups <- c("FL", "PA", "SSF", "LSF", "Env")

variates_list <- lapply(seq_along(groups), function(i) {
  df <- as.data.frame(gca_out$canonical_variates[[i]])
  
  # give component names
  colnames(df) <- c("GCA1", "GCA2", "GCA3", "GCA4", "GCA5")
  
  # add block label
  df$Group <- groups[i]
  
  # create sample index (1–678)
  df$Sample_num <- seq_len(nrow(df))
  
  df
})

# combine all blocks
variates_df <- do.call(rbind, variates_list)

## add back sample names
M_FL_scaled$Sample_num <- seq_len(nrow(M_FL_scaled))
variates_df <- left_join(variates_df, M_FL_scaled, by = "Sample_num")
variates_df$Sample_num<-rownames(M_FL_scaled)
M_FL_scaled<-M_FL_scaled %>% select(-Sample_num)

#add cat meta
cat_meta<- FL_meta %>% select(Sample_num, YEAR, SEASON, lake,new_depths_mdh2,layer_name2 )
variates_df<- variates_df %>% left_join(cat_meta, by = "Sample_num")

head(variates_df)
```
```{r}
block_contrib <- variates_df %>%
  group_by(Group) %>%
  summarise(
    GCA1 = mean(abs(GCA1), na.rm = TRUE),
    GCA2 = mean(abs(GCA2), na.rm = TRUE),
    GCA3 = mean(abs(GCA3), na.rm = TRUE),
    GCA4 = mean(abs(GCA4), na.rm = TRUE),
    GCA5 = mean(abs(GCA5), na.rm = TRUE),
    n    = n(),
    .groups = "drop"
  )

block_contrib

p_blocks<- block_contrib %>%
  pivot_longer(
    cols = starts_with("GCA"),
    names_to = "Axis",
    values_to = "MeanAbsLoading"
  ) %>%
  filter(Axis %in% c("GCA1", "GCA2","GCA3","GCA4","GCA5")) %>%
  mutate(Group=factor(Group, levels=c("Env","FL","SSF","PA","LSF"))) %>% 
  ggplot(aes(x = Group, y = MeanAbsLoading, fill = Axis)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "GCA1" = "#0072B2",  # blue
    "GCA2" = "#E69F00",  # amber
    "GCA3" = "#009E73",  # teal-green
    "GCA4" = "#CC79A7",   # pink-purple
    "GCA5" = "darkred"
  )) +
  theme_classic() +
  labs(
    y = "Mean |Canonical Variate|",
    title = expression(bold("A") * "")
  ) +
  guides(fill = guide_legend(ncol = 4))+
  theme(
    legend.position = c(0.4, 0.92),        # x, y in 0-1 plot coordinates
    legend.background = element_rect(fill = "white", color = "grey80"),
    legend.margin = margin(4, 6, 4, 6)
  )

#ggsave("../output/plots/Multiple/GCA_block_contributions.png",p, width = 5, height = 4, units = "in", dpi = 600)

p_blocks
```

#### Canonical vectors  
```{r}
#Extract env canonical vectors (block 5)
env_vec <- as.data.frame(gca_out$canonical_vectors[[5]])

# Name the columns for the GCA components
colnames(env_vec) <- paste0("GCA", seq_len(ncol(env_vec)))

# Add variable names as a column
env_vec$variable <- colnames(M_FL_scaled)

# Reshape to long format for faceting
env_long <- env_vec %>%
  pivot_longer(cols = starts_with("GCA"),
               names_to = "Component",
               values_to = "Loading")

env_12 <- env_long %>% 
  filter(Component %in% c("GCA1", "GCA2","GCA3","GCA4","GCA5")) %>%  #
  group_by(Component) %>%
  mutate(
    abs_loading = abs(Loading),
    highlight = abs_loading %in% sort(unique(abs_loading), decreasing = TRUE)[1:3]
  ) %>%
  ungroup()

## we need to do this becayse here light is not in logit transformation
meta_names<-read.csv("/Volumes/mlcoleman/project/greatlakes/Env_Metadata/meta_names.csv")
param_order2 <- sub("estimated_light_logit", "estimated_light_pcnt", param_order)
new_order2 <- meta_names$new[match(param_order2, meta_names$original)]

keep2=c("Alkalinity..Total.as.CaCO3_mg.l","Chlorophyll.a_ug.l",
       "Nitrogen..Total.Oxidized_mg.l","CTD_TEMP", "estimated_light_pcnt","Phosphorus..Total.as.P_ug.l",
      "Silica..Dissolved.as.Si_mg.l","Turbidity_NTU", "abs_depth_m", "STN_DEPTH_M", 
      "sbeox0Mg.L", "airT_14d_mean")

env_12B <- env_12 %>%
  filter(variable %in% keep2) %>% 
  left_join(meta_names, by = c("variable" = "original")) %>%
  mutate(variable = coalesce(new, variable))  %>% 
  mutate(variable= factor(variable, levels = new_order2))

p_loadings_env <- ggplot(env_12B, 
       aes(x = variable, y = Loading, 
           fill = highlight, color = highlight, linewidth = highlight)) +
  geom_col() +
  scale_fill_manual(values = c("FALSE" = alpha("steelblue", 0.4), 
                               "TRUE"  = alpha("steelblue", 1.0))) +
  scale_color_manual(values = c("FALSE" = alpha("steelblue", 0.4), 
                                "TRUE"  = "steelblue4")) +
  scale_linewidth_manual(values = c("FALSE" = 0.2, 
                                    "TRUE"  = 0.8)) +
  coord_flip() +
  facet_wrap(Component ~., ncol = 5) +
  labs(
    title = expression(bold("B") * ""),
    x = "Environmental Variable",
    y = "Loading") +
  theme_minimal(base_size = 13) +
  guides(fill = "none", color = "none", linewidth = "none")

#ggsave("/output/plots/Multiple/GCA_env_loadingss.png",p, width = 4, height = 6, units = "in", dpi = 600)

p_loadings_env
```

#### odinations -- fix light
```{r}
clean_variates<-subset(variates_df,Group != "Env")
clean_variates
```

```{r}
#the arrows
top_vars <- env_12 %>%
  dplyr::filter(highlight) %>%
  dplyr::distinct(variable)

env_arrows <- env_vec %>%
  dplyr::semi_join(top_vars, by = "variable") %>%
  dplyr::select(variable, GCA1, GCA2)

#some scaling
# rough scale based on spread of scores
range_x <- range(variates_df$GCA1, na.rm = TRUE)
range_y <- range(variates_df$GCA2, na.rm = TRUE)
arrow_scale <- 2 * max(diff(range_x), diff(range_y))   # tweak factor if needed

env_arrows_scaled <- env_arrows %>%
  mutate(
    xend = GCA1 * arrow_scale,
    yend = GCA2 * arrow_scale
  ) %>% left_join(meta_names, by = c("variable" = "original")) %>%
  mutate(variable = coalesce(new, variable)) 

env_arrows_scaled <- env_arrows_scaled %>%
  mutate(vec_length = sqrt(GCA1^2 + GCA2^2)) %>%
  filter(vec_length > quantile(vec_length, 0.2))  # keep above the X percentile

env_arrows_scaled 
```



```{r}
p_lake<- ggplot(clean_variates, aes(GCA1, GCA2, color = lake)) +
  scale_color_manual(values=lake_colors)+
  geom_point(alpha = 0.8,size=3) +
  geom_vline(xintercept = 0, color = "gray60", linewidth = 0.4,linetype = "dashed") +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.4,linetype = "dashed") +
  theme_bw()+
  facet_wrap(.~Group)+
  # environmental arrows (from origin)
  geom_segment(data = env_arrows_scaled, #subset(env_arrows_scaled, variable %in% params_wanted)
               aes(x = 0, y = 0, xend = xend, yend = yend),
               inherit.aes = FALSE,
               arrow = arrow(length = unit(0.2, "cm")),
               linewidth = 0.8,
               color = "black") +
  
  # labels for env vectors
  ggrepel::geom_label_repel(
  data = env_arrows_scaled,
  aes(x = xend , y = yend , label = variable),
  inherit.aes   = FALSE,
  color         = "white",
  fill      = "#2C2C2C",
  fontface  = "bold",
  alpha = 0.7,
  size          = 5,
  box.padding   = 0.5,
  point.padding = 0.3,
  segment.color = "black",
  segment.size  = 0.3,
  max.overlaps  = Inf,
  show.legend   = FALSE) +
  labs(color="Lake",title=expression(bold("B") * ""),
       x = "GCA1 (16.3%)",
  y = "GCA2 (12.0%)")

#ggsave("../output/plots/Multiple/GCA_lake.png", p, width = 10, height = 8, units = "in", dpi = 600)

p_lake
```
```{r}
p_strata<-ggplot(clean_variates, aes(GCA1, GCA2, color = new_depths_mdh2)) +
  scale_color_manual(values=depth2_colors)+
  #scale_shape_manual(values=lake_shapes)+
  geom_point(alpha = 0.8,size=3) +
  geom_vline(xintercept = 0, color = "gray60", linewidth = 0.4,linetype = "dashed") +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.4,linetype = "dashed") +
  theme_bw()+
  facet_wrap(.~Group)+
  # environmental arrows (from origin)
  geom_segment(data = env_arrows_scaled,
               aes(x = 0, y = 0, xend = xend, yend = yend),
               inherit.aes = FALSE,
               arrow = arrow(length = unit(0.2, "cm")),
               linewidth = 0.8,
               color = "black") +
  
  # labels for env vectors
  ggrepel::geom_label_repel(
  data = env_arrows_scaled,
  aes(x = xend , y = yend , label = variable),
  inherit.aes   = FALSE,
  color         = "white",
  fill      = "#2C2C2C",
  fontface  = "bold",
  alpha = 0.7,
  size          = 5,
  box.padding   = 0.5,
  point.padding = 0.3,
  segment.color = "black",
  segment.size  = 0.3,
  max.overlaps  = Inf,
  show.legend   = FALSE) +
  labs(color="Water Layer",
       title=expression(bold("A") * ""),
       x = "GCA1 (16.3%)",y = "GCA2 (12.0%)")

#ggsave("../output/plots/Multiple/GCA_layer.png",p, width = 10, height = 8, units = "in", dpi = 600)

p_strata
```

### GCA summary figure

```{r}
gca_out1 <- cowplot::plot_grid(
  p_strata, p_lake,
  ncol = 2, nrow = 1
)

ggsave("output/plots/ALL/gca_ord.pdf", gca_out1, width = 14, height = 8, units = "in",dpi=600)
gca_out1

gca_out2 <- cowplot::plot_grid(
  p_blocks, p_loadings_env,
  ncol = 1, nrow = 2
)

ggsave("output/plots/ALL/gca_summary.pdf", gca_out2, width = 10, height = 8, units = "in",dpi=600)
gca_out2
```


## Heart Plots
```{r}
#install.packages("ggnewscale") 
library(ggnewscale)
```

```{r}
main_stations=c("SU18","SU17M","SU15","SU06","SU08M","SU01M",
                "MI41M","MI27M","MI18M",
                "HU54M","HU45M","HU15M",
                "ER91M","ER78M","ER15M",
                "ON33M","ON41","ON55M")

# build the universal station_year levels from all metadata
all_station_year_df <- bind_rows(
  FL_meta  %>% select(STATION_ID, YEAR),
  PA_meta  %>% select(STATION_ID, YEAR),
  SSF_meta %>% select(STATION_ID, YEAR),
  LSF_meta %>% select(STATION_ID, YEAR)
) %>%
  distinct() %>%
  filter(STATION_ID %in% main_stations) %>%
  mutate(YEAR = as.numeric(as.character(YEAR))) %>%  # <-- make sure it is numeric
  arrange(factor(STATION_ID, levels = main_stations), YEAR) %>%
  mutate(station_year = paste(STATION_ID, YEAR, sep = "_"))

all_station_year <- all_station_year_df %>% pull(station_year)
all_station_year
```

```{r}
get_heart2 <- function(plot_title, name, df, meta, 
                       value_num = 0.05, 
                       shared_x_df = all_station_year_df,
                       show_x = TRUE){
  
  layer <- unique(meta$new_depths_mdh2)[1]
  
  if (name == "FL"){
    list_colors <- FL_colors
  } else if (name == "PA"){
    list_colors <- PA_colors
  } else if (name == "SSF"){
    list_colors <- SSF_colors
  } else if (name == "LSF"){
    list_colors <- LSF_colors
  } else {
    return("check inputs")
  }
  
  # reshape
  df$Sample_num <- rownames(df) 
  df <- df %>% rename_with(~ paste0(name, "_", .), where(is.numeric))
  
  df_long <- df %>%
    pivot_longer(cols = starts_with(name), names_to = "topic", values_to = "value")
  
  df_long <- df_long %>% 
    select(Sample_num, topic, value) %>% 
    filter((Sample_num %in% meta$Sample_num) & value > value_num) %>% 
    left_join(meta, "Sample_num") %>% 
    dplyr::select(STATION_ID, YEAR, topic, value) %>% 
    mutate(STATION_ID = factor(STATION_ID, levels = main_stations),
           YEAR = factor(YEAR, levels = 2012:2019),
           station_year = paste(STATION_ID, YEAR, sep = "_"),
           topic_num = as.numeric(gsub(".*_", "", topic)))
  
  # topic order
  topic_levels <- df_long %>%
    distinct(topic, topic_num) %>%
    arrange(topic_num) %>%
    pull(topic)
  
  # station_year order from shared_x_df
  ordered_station_year <- shared_x_df %>% pull(station_year)
  
  # station breaks from shared_x_df
  station_breaks <- shared_x_df %>%
    group_by(STATION_ID) %>%
    filter(YEAR == max(YEAR)) %>%
    ungroup() %>%
    mutate(x_pos = match(station_year, ordered_station_year) + 0.5)
  
  # apply factor levels
  df_long <- df_long %>%
    mutate(station_year = factor(station_year, levels = ordered_station_year),
           topic = factor(topic, levels = topic_levels))

  # lake color labels
  lake_labels <- df_long %>%
    distinct(station_year, STATION_ID) %>%
    mutate(lake_code = substr(as.character(STATION_ID), 1, 2),
           x = match(station_year, levels(df_long$station_year)))

  p <- ggplot(df_long, aes(x = station_year, y = topic, size = value, fill = topic)) +
    geom_vline(data = station_breaks, aes(xintercept = x_pos), 
               linetype = "dashed", color = "grey40") +
    geom_tile(aes(height = value), col = "transparent") +
    scale_size_continuous(range = c(0.1, 5)) +
    scale_fill_manual(values = list_colors) +
    scale_x_discrete(drop = FALSE) +  # <-- keep all shared x levels
    theme_bw() +
    labs(x = "", y = "Subcommunity", size = "Value", title = plot_title) +
    theme(
      plot.margin = if (show_x) margin(1, 5, 1, 5) else margin(1, 5, -5, 5),
      plot.title = element_text(size = 6, face = "bold"), 
      panel.grid = element_blank(),
      axis.text.x = if (show_x) element_text(angle = 90, hjust = 1, size = 5) else element_blank(),
      axis.ticks.x = if (show_x) element_line() else element_blank(),
      legend.position = "right") +
    guides(fill = "none", size = "none") +
    ggnewscale::new_scale_fill() +
    geom_tile(data = lake_labels,
              aes(x = x, y = -0.5, fill = lake_code), 
              width = 1, height = 0.5,
              inherit.aes = FALSE) +
    scale_fill_manual(values = lake_colors) +
    guides(fill = "none")
  
  return(p)
}
  
wanted<-subset(PA_meta, new_depths_mdh2 == "Strat_EPI" & SEASON == "Summer")
get_heart2("PA Prok Summer Strat_EPI (p>0.05)", "PA", W_PA, wanted, 
           value_num = 0.1, shared_x_df = all_station_year_df)
```
###strat EPI
```{r}
"only show subcommunities contributing more than twice their fair share, so value_num =2/k"

#FL
wanted <- subset(FL_meta,  new_depths_mdh2 %in% c("Strat_EPI","Iso_EPI") & SEASON == "Summer")
FL_epi  <- get_heart2("FL Prok",    "FL",  W_FL,  wanted, value_num = 3/ncol(W_FL), show_x = FALSE)

#PA
wanted <- subset(PA_meta,  new_depths_mdh2 %in% c("Strat_EPI","Iso_EPI") & SEASON == "Summer")
PA_epi  <- get_heart2("PA Prok",    "PA",  W_PA,  wanted, value_num = 3/ncol(W_PA), show_x = FALSE)

#SSF
wanted <- subset(SSF_meta, new_depths_mdh2 %in% c("Strat_EPI","Iso_EPI") & SEASON == "Summer")
SSF_epi <- get_heart2("SSF Chloro", "SSF", W_SSF, wanted, value_num = 3/ncol(W_SSF), show_x = FALSE)

#LSF
wanted <- subset(LSF_meta, new_depths_mdh2 %in% c("Strat_EPI","Iso_EPI") & SEASON == "Summer")
LSF_epi <- get_heart2("LSF Chloro", "LSF", W_LSF, wanted, value_num = 3/ncol(W_LSF), show_x = TRUE)


##save
figure <-  ggpubr::ggarrange(FL_epi, PA_epi, SSF_epi,LSF_epi, ncol = 1, nrow = 4,
                             align = "v",heights = c(1, 1, 1, 1.5),hjust = -0.1)

figure<-annotate_figure(figure,
                top = text_grob("Strat Epilimnion", color = "black", face = "bold", size = 14))

ggsave(file=paste("output/plots/ALL/strat_epi.png",sep=''), figure,
       width = 10, height = 12, units = "in",dpi=1000)

figure
```


### Cold water
```{r}
###cold
#FL
wanted <- subset(FL_meta, new_depths_mdh2 %in% c("Strat_HYP","Iso_Bot","Iso_Top","Iso_Mid","InvS_EPI","InvS_HYP"))
FL_cold  <- get_heart2("FL Prok",    "FL",  W_FL,  wanted, value_num = 3/ncol(W_FL),  show_x = FALSE)

#PA
wanted <- subset(PA_meta, new_depths_mdh2 %in% c("Strat_HYP","Iso_Bot","Iso_Top","Iso_Mid","InvS_EPI","InvS_HYP"))
PA_cold  <- get_heart2("PA Prok",    "PA",  W_PA,  wanted, value_num = 3/ncol(W_PA),  show_x = FALSE)

#SSF
wanted <- subset(SSF_meta, new_depths_mdh2 %in% c("Strat_HYP","Iso_Bot","Iso_Top","Iso_Mid","InvS_EPI","InvS_HYP"))
SSF_cold <- get_heart2("SSF Chloro", "SSF", W_SSF, wanted, value_num = 3/ncol(W_SSF), show_x = FALSE)

#LSF
wanted <- subset(LSF_meta, new_depths_mdh2 %in% c("Strat_HYP","Iso_Bot","Iso_Top","Iso_Mid","InvS_EPI","InvS_HYP"))
LSF_cold <- get_heart2("LSF Chloro", "LSF", W_LSF, wanted, value_num = 3/ncol(W_LSF), show_x = TRUE)

##save
figure <- plot_grid(FL_cold, PA_cold, SSF_cold, LSF_cold,
                    ncol = 1, rel_heights = c(1, 1, 1, 1.5))
figure <- ggpubr::annotate_figure(figure,
                top = ggpubr::text_grob("Cold Waters", color = "black", face = "bold", size = 14))
ggsave(file = "output/plots/ALL/cold_water.png", figure,
       width = 10, height = 12, units = "in", dpi = 1000)
figure
```



